---
title: "Indicator Variables"
author: "Cody Frisby"
date: "March 10, 2016"
output: 
  html_document: 
    keep_md: yes
---

Salary example shared in class.  

```{r, dpi=200, fig.align='center', cache=TRUE, message=FALSE, warning=FALSE}
library(xlsx) # read excel files
sal <- read.xlsx("~/Documents/MATH3710/salaryIV/Salary.xlsx", 
                 sheetIndex = 1)
sal <- sal[1:45,] # remove NA lines from bottom of sal
```

We need to recode some of these vars.  Here's one, simple, way to do it in R.

```{r}
sal$bachelors <- 0 #create a new var with zeros
sal$grad_deg <- 0 # ditto
# we need statement to fill in our new vars with 1s
sal[sal$Education == "BS", "bachelors"] <- 1
sal[sal$Education == "BS+", "grad_deg"] <- 1
# now one for manager
sal$mgr <- 0
sal[sal$Manager == "Yes", "mgr"] <- 1
sal$EM1 <- sal$bachelors*sal$mgr
sal$EM2 <- sal$grad_deg*sal$mgr
head(sal) # look at the data
```

Now, for ease, I'm going to subset our data table. Sal2 is a new data frame that contains the variables we are interested in.

```{r}
sal2 <- sal[,c(1:2,6:8)]
head(sal2)
```

### Box plot of Salary by Education level

```{r, fig.align='center', dpi=200, message=F, warning=F}
library(ggplot2)
bx <- ggplot(sal, aes(Education, Salary))
bx <- bx + geom_boxplot(fill = "lightblue") + geom_jitter() + theme_bw()
bx
```

### Box plot of Salary by Manager

```{r, fig.align='center', dpi=200, message=F, warning=F}
library(ggplot2)
bx <- ggplot(sal, aes(Manager, Salary))
bx <- bx + geom_boxplot(fill = "lightblue") + geom_jitter() + theme_bw()
bx
```


###More Plotting, scatter plots with color and shape as dimensions

```{r, dpi=200, fig.align='center', cache=TRUE, message=FALSE, warning=FALSE}
g <- ggplot(data = sal, aes(x=Experience, y=Salary))
g <- g + geom_point(aes(shape=Education, colour=Manager), size = 3)
g
```

Now, let's fit the model, using indicator variables

```{r}
fit <- lm(Salary ~ ., sal2) #data frame with only the vars we want
summary(fit)
```

### Residual Diagnostics

```{r, dpi=200, fig.align='center'}
par(mfrow=c(1,2))
plot(fit, which = c(1,2))
```

By the looks of it, we have an issue with "clustering" possibly due to th catagories.  There may be an interaction we are not accounting for in our model. 

### Model with interaction

```{r}
fit2 <- lm(Salary ~ . + grad_deg*mgr+bachelors*mgr, sal2)
summary(fit2)
```

### Model Diagnostics  

```{r, dpi=200, fig.align='center'}
par(mfrow=c(1,2))
plot(fit2, which = c(1,2))
```

Looking much better now.

