---
title: "Chapter 3 Homework"
author: "Cody Frisby"
date: "September 23, 2016"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, dpi=200, fig.align = 'center',
                      warning = F, message = F)
```

## 3.1  

```{r}
# 3.1 - 3.10 is the assigment.  From Statistical Quality Control
# sixth edition. 
df <- read.csv("~/Documents/STAT4600/hw/ch03/3-1.csv")
x <- df[,2]
```

The mean (sample average) for these data is `r mean(x)`.  The sample standard deviation is calculated by $$\hat \sigma = \sqrt{\frac{\sum_{i=1}^{n}(x_i - \bar x)^2}{n-1}}$$ where $\bar x$ is the sample mean.  For our sample this is equal to $`r sd(x)`$.  

## 3.2  

```{r}
x <- read.csv("~/Documents/STAT4600/hw/ch03/3-2.csv")
x <- x[,2]
```

The data are $$`r x`$$
The mean for these data is `r mean(x)`.  
The standard deviation for these data is $`r sd(x)`$.  

## 3.3  

```{r}
df <- read.csv("~/Documents/STAT4600/hw/ch03/3-3.csv")
x <- df[,2]
```

The data are $$`r x`$$
The mean for these data is `r mean(x)`.  
The standard deviation for these data is $`r sd(x)`$. 

## 3.4  

First I'd like to order the data $$`r sort(x)`$$  Now we'd just find the middle value to estimate the median with, the median is equal to `r median(x)`.  If our sample size remains the same, 9, then the largest value could be as large as infinity (or more practically, as high as our measurement device or furnace could go) without changing the median.  

## 3.5  

```{r}
df <- read.csv("~/Documents/STAT4600/hw/ch03/3-5.csv")
x <- df[,2]
```

The data are $$`r x`$$
The mean for these data is `r mean(x)`.  
The standard deviation for these data is $`r sd(x)`$. 

## 3.6  

```{r}
df <- read.csv("~/Documents/STAT4600/hw/ch03/3-6.csv")
x <- df[,2]
```

The mean for these data is `r mean(x)`.  
The standard deviation for these data is $`r sd(x)`$

A histogram is below:  

```{r, fig.width=5}
hist(x, col="green", xlab="Time to Failure", main="")
```

And a stem and leaf plot is below:  

```{r, comment=""}
# stem and leaf plot
stem(x)
```

## 3.7  

Histogram is produced here:  

```{r, fig.width=5}
df <- read.csv("~/Documents/STAT4600/hw/ch03/3-7.csv")
x <- df[,2]
h <- hist(x, col="green", xlab="Yield", main="", freq= F, plot=F)
m <- h$counts/h$density
d <- density(x)
d$y <- d$y * m[1]
plot(h, col="green", xlab="Yield", main="")
lines(d, col="blue", lwd=2)
# add a normal curve
xfit<-seq(min(x)-2,max(x)+1,length=length(d$x)) 
yfit<-dnorm(xfit,mean=mean(x),sd=sd(x)) 
yfit <- yfit*diff(h$mids[1:2])*length(x) 
lines(xfit, yfit, col="red", lwd=2)
# does the data follow a t distribution?
```

Not sure which distribution this histogram might look like.  It definitely doesn't look like a normal distribution (red curve).  I've added a density curve (blue) to the plot as a reference.  The best fit for our data may not be one which is covered in chapter 3. How about looking at an emperical distribution plot:  

```{r}
plot(ecdf(x), verticals = TRUE, pch="")
curve(punif(x, min(x), max(x)), col="red", lty=3, add=TRUE)
curve(pt(x-mean(x), length(x)), col="blue", lty=3, add=TRUE)
curve(pnorm(x, mean(x), sd(x)), col="black", lty=3, add=TRUE)
```


## 3.8  

First, construct and display a stem and leaf plot:  

```{r, comment = ""}
df <- read.csv("~/Documents/STAT4600/hw/ch03/3-8.csv")
x <- df[,2]
stem(x)
```

And a plost of thei histogram with a fitted normal and density curve: 

```{r, fig.width=5}
h <- hist(x, plot=F)
m <- h$counts/h$density
d <- density(x)
d$y <- d$y * m[1]
plot(h, col="green", xlab="Viscosity", main="")
lines(d, col="blue", lwd=2)
# add a normal curve
xfit<-seq(min(x)-2,max(x)+1,length=length(d$x)) 
yfit<-dnorm(xfit,mean=mean(x),sd=sd(x)) 
yfit <- yfit*diff(h$mids[1:2])*length(x) 
lines(xfit, yfit, col="red", lwd=2)
# t dist
xfit<-seq(min(x)-2,max(x)+1,length=length(d$x)) 
yfit<-dt(xfit-mean(x),df = length(xfit) -1) 
yfit <- yfit*diff(h$mids[1:2])*length(x) 
lines(xfit, yfit, col="black", lwd=2)
plot(ecdf(x), verticals = TRUE, pch="")
curve(pnorm(x, mean(x), sd(x)), col="red", lty=3, add=TRUE)
curve(pt(x-mean(x), length(x)-1), col="blue", lty=3, add=TRUE)
```

Interestingly, unlike the prior histogram, the normal and density curves appear to be very close (red) with the t distribution in black as well.  Also, the blue line (t distribution) follows the cdf plot of the data very closely.  Assuming the data follows the t distribution appears a safe call.  

The above stem and leaf plot 

```{r, eval=F}
# the source data are in an excel file that has differing rows and 
# columns.  This is how I am importing it
df <- read.clipboard()
write.csv(df, "hw/hw06/8-26.csv")
```

