---
title: "Life Expectancy"
author: "Cody Frisby"
date: "March 21, 2016"
output: pdf_document
---

```{r, echo=FALSE, warning=F, message=F, cache=TRUE}
# read in the data from the excel file
life <- xlsx::read.xlsx("~/Documents/MATH3710/ProblemSets/problem6/LifeExpectancy.xlsx", sheetIndex = 1)
life$africa <- 0
life$oecd <- 0
life[life$Group == "africa", "africa"] <- 1
life[life$Group == "oecd", "oecd"] <- 1
```

### 1)  MLR and Variables to include  

Here we plot the log of PPGDP by LifeExp with color/shape overlay of Group.  

```{r, echo=F, dpi=200, fig.align='center'}
library(ggplot2)
g <- ggplot(data = life, aes(x=log(PPGDP), y=LifeExp))
g <- g + geom_point(aes(shape = Group, colour = Group), size = 4) + theme_bw()
g
```

Now to look at box plots LifeExp by Group.

```{r, echo=F, dpi=200, fig.align='center'}
b <- ggplot(life, aes(Group, LifeExp))
b <- b + geom_boxplot(fill = "lightblue") + geom_jitter() + theme_bw()
b
# below code creates a plot of predicted by observed as a way 
# to visually see how "good" our model is.
fit.all <- lm(LifeExp ~ log(PPGDP)*africa + log(PPGDP)*oecd, data = life)
mod <- lm(life$LifeExp ~ fit.all$fitted.values)
rout <- list(paste('Fitted model: ', round(coef(mod)[1], 4), ' + ',
                   round(coef(mod)[2], 3), ' x', sep = ''),
             paste('R^2 == ', round(summary(mod)[['r.squared']], 3),
                   sep = ''))
#rout
df <- as.data.frame(cbind(x = fit.all$fitted.values, y = life$LifeExp))

fitplot <- ggplot(df, aes(x, y)) +
  geom_smooth(method = 'lm') + geom_point() + xlab("Predicted") + 
  ylab("Observed") +
# need to change these x and y values below
  #geom_text(aes(x = 52, y = 80, label = rout[[1]]), hjust = 0) +
  geom_text(aes(x = 52, y = 78, label = rout[[2]]), hjust = 0, 
            parse = TRUE)
```

It appears that the indicator variables should be included.  Group appears to have an influence on the life expectancy.  There doesn't seem to be immediate need to include interaction terms in our model.  Group and log(PPGDP) appear to be related but there doesn't appear to be an interaction between them.  Also of note, group may not be needed at all, except **africa**, since most of the explanatory of **LifeExp** appears to be explained by **log(PPGDP)**.    

### 2) Model  

```{r, echo=F}
fit2 <- lm(LifeExp ~ log(PPGDP)*Group, life) # not needed, same as fit.all
f <- formula(LifeExp~log(PPGDP)+africa+oecd+log(PPGDP)*africa+log(PPGDP)*oecd) 
# less typing when we use this ^.
fit <- lm(LifeExp ~ log(PPGDP)+africa+oecd, data = life)
# no interactions ^. 
```

Note: $X_1 = log(PPGDP)$, $X_2 = 1$, if *Group* = **africa**, $X_3 = 1$, if *Group* = **oecd**.  The model is $$y = \beta_0 + \beta_1X_1 + \beta_2X_2 + \beta_3X_3 + \beta_4X_1X_2 + \beta_5X_1X_3 + \varepsilon$$ where $\beta_0$ is the expected life expectancy with no PPGDP and group equal to **other**, $\beta_1$ is the expected change in life expectancy given **log(PPDGP)** when country group is **other**, $\beta_2$ is the difference between an **africa** country and **other**, and $\beta_3$ is the difference between an **oecd** country and **other**, $\beta_4$ is the interaction between **log(PPDGP)** and *group* **africa**, $\beta_5$ is the interaction between **log(PPDGP)** and *group* **oecd**.  These last two, $\beta_4$ &  $\beta_5$ are simply additive, meaning how much additional or less *LifeExp* can we expect compared to *group* **other** when country is *Africa* or *oecd*, when considering **log(PPGDP)**.  Here, we assume the errors from the above model are *normal* (0, $\sigma^2$) and we assume a linear relationship between the predictors and the response **LifeExp**.  This model will allow us to answer whether or not life expectancy is linked to PPGDG as well as quantify whether or not **OECD** countries have longer life expediencies. 

### 3)  Fit Model  

```{r, echo=F}
b <- coef(fit.all)
```

Our fitted models' equation (from above) is $$LifeExp = `r b[1]` + `r b[2]`X_1  `r b[3]`X_2 + `r b[4]`X_3 + `r b[5]`X_1X_2  `r b[6]`X_1X_3$$ and here I display a scatter plot with our fitted models.  

```{r, echo=F, dpi=200, fig.align='center'}
# Slopes and intercepts from linear model
library(plyr)
coefs <- ddply(life, .(Group), function(df) {
  m <- lm(LifeExp ~ log(PPGDP), data = df)
  data.frame(a = coef(m)[1], b = coef(m)[2])
})
g <- ggplot(data = life, aes(x=log(PPGDP), y=LifeExp))
g <- g + geom_point(aes(shape = Group, colour = Group), size = 4) + theme_bw()
g <- g + geom_smooth(aes(group=Group), method = "lm", se=F, fullrange=F)
#g <- g + geom_abline(data=coefs, aes(intercept=a, slope=b))
g
```

Where the top line is **oecd**, middle line is **other**, and the bottom is **africa**.  

### 4)  Reduced Model  

```{r, echo=F}
fit.red <- lm(LifeExp ~ log(PPGDP) + africa, life)
b0 <- coef(fit.red)
```

The reduced model will be $$LifeExp = `r b0[1]` + `r b0[2]`X_1  `r b0[3]`X_2$$ where, as before, $X_1$ is *log(PPGDP)* and $X_2 = 1$, if *Group* = *africa*. Here, we've removed all interactions and are now simply comparing groups *other* and *africa*.  Displaying a plot below 

```{r, echo=F, dpi=200, fig.align='center'}
df <- data.frame(b0 = c(b0[1], b0[1]+b0[3]), b1 = c(b0[2],b0[2]))
g <- ggplot(data = life, aes(x=log(PPGDP), y=LifeExp))
g <- g + geom_point(aes(shape = Group, colour = Group), size = 4) + theme_bw()
g <- g + geom_abline(data=df, aes(intercept=b0, slope=b1))
g
```

where the top line is **other** and the bottom is **africa**.  

### 5) Model Adequacy  

Here, I take a look at some residual diagnostics for the above, reduced, model.  

```{r, echo=F, dpi=200, fig.align='center'}
par(mfrow=c(1,2))
plot(fit.red, which = c(1,2))
```

Of concern is the equality of variances assumption and the normality assumption.  They aren't perfect, they don't need to be though, and we have transformed one of the predictor variables and this is often the case when we do a transformation.  I think they are close enough and we will proceed with the above model.  The reduced models $R^2$ is `r summary(fit.red)$r.squared`.  This means that we can explain approximately 75% of the variation in life expectancy by knowing whether or not the country is from Africa and knowing the PPGDP of the country.  

### 6)  F test for the removal of covariates  

```{r, echo=F}
sse.fm <- anova(fit.all)[[2]][6]
sse.rm <- anova(fit.red)[[2]][3]
mse.rm <- sse.rm/anova(fit.red)[[1]][3]
f.stat <- (sse.rm - sse.fm)/(mse.rm)
F0 <- qf(0.95, 3, 196)
f.p <- 1 - pf(f.stat, 3, 196)
```

$$F = \frac{\frac{SSE_{RM} - SSE_{FM}}{n - k}}{MSE_{RM}}$$
$$`r f.stat` = \frac{\frac{`r sse.rm` - `r sse.fm`}{3}}{`r mse.rm`}$$  The critical value for $F_{3,196:\alpha=0.05}$ is `r F0`.  So, we conclude that we have not adversely effected the model by removing the terms that we have removed.  

```{r, echo=F}
knitr::kable(anova(fit.red))
```

Above, I display an ANOVA table of our reduced model.  Included are *F* tests for the terms in the model.  We can concluded that each of the included terms significantly effect *life expectancy*.  
  
Below, I display 95% confidence intervals for each of our reduced model terms.  
*Intercept*, which is interpretted as the expected *life expectancy* of countries in **Group** *oecd* or *other*, *on average*.  *log(PPGDP)* is interpreted as the expected change in *life expectancy* per unit change in PPGDP, *on average*. And finally, **africa**, is interpreted as the expected life expectancy when comparted to *other* countries, *on average*.  

```{r, echo=F}
knitr::kable(confint(fit.red))
```

### 7)  Conclusions  

We can answer YES to the question of whether or not a country's well being is linked with life expectancy.  The inverval above, [`r confint(fit.red)[2,]`], quantifies the effect that PPGDP has on life expectancy.  For model assumptions to work we have taken the log of this value, but this interval does not include zero nor is it very close to zero.  On average, we except that the log(PPGDP) has an effect on life expectancy by at least 2.8.  
  
  **OECD** countries will, on average, have higher life expectancies than **africa** countries but not nesecarrily **other** countries.  If we consider **other** vs. **africa** where **other** is including **oecd** countries, then the interval displayed above for the line *africa* quantifies this difference for us.  We can expect **other** countries to have a life expectancy of at least 8.6 points higher than *Africa* countries. 
  
### R code:  

```{r, eval=F}
# read in the data from the excel file
life <- xlsx::read.xlsx
("~/Documents/MATH3710/ProblemSets/problem6/LifeExpectancy.xlsx", sheetIndex = 1)
# create indicator variables
life$africa <- 0
life$oecd <- 0
life[life$Group == "africa", "africa"] <- 1
life[life$Group == "oecd", "oecd"] <- 1
# plot with shape/color overlay
library(ggplot2)
g <- ggplot(data = life, aes(x=log(PPGDP), y=LifeExp))
g <- g + geom_point(aes(shape = Group, colour = Group), size = 4) + theme_bw()
g
b <- ggplot(life, aes(Group, LifeExp))
b <- b + geom_boxplot(fill = "lightblue") + geom_jitter() + theme_bw()
b
# below code creates a plot of predicted by observed as a way 
# to visually see how "good" our model is.
fit.all <- lm(LifeExp ~ log(PPGDP)*africa + log(PPGDP)*oecd, data = life)
mod <- lm(life$LifeExp ~ fit.all$fitted.values)
rout <- list(paste('Fitted model: ', round(coef(mod)[1], 4), ' + ',
                   round(coef(mod)[2], 3), ' x', sep = ''),
             paste('R^2 == ', round(summary(mod)[['r.squared']], 3),
                   sep = ''))
#rout
df <- as.data.frame(cbind(x = fit.all$fitted.values, y = life$LifeExp))

fitplot <- ggplot(df, aes(x, y)) +
  geom_smooth(method = 'lm') + geom_point() + xlab("Predicted") + 
  ylab("Observed") +
# need to change these x and y values below
  #geom_text(aes(x = 52, y = 80, label = rout[[1]]), hjust = 0) +
  geom_text(aes(x = 52, y = 78, label = rout[[2]]), hjust = 0, 
            parse = TRUE)
fit2 <- lm(LifeExp ~ log(PPGDP)*Group, life) # not needed, same as fit.all
f <- formula(LifeExp~log(PPGDP)+africa+oecd+log(PPGDP)*africa+log(PPGDP)*oecd) 
# less typing when we use this ^.
fit <- lm(LifeExp ~ log(PPGDP)+africa+oecd, data = life)
# no interactions ^. 
b <- coef(fit.all) # model coefficients
# Slopes and intercepts from linear model
library(plyr)
coefs <- ddply(life, .(Group), function(df) {
  m <- lm(LifeExp ~ log(PPGDP), data = df)
  data.frame(a = coef(m)[1], b = coef(m)[2])
})
g <- ggplot(data = life, aes(x=log(PPGDP), y=LifeExp))
g <- g + geom_point(aes(shape = Group, colour = Group), size = 4) + theme_bw()
g <- g + geom_smooth(aes(group=Group), method = "lm", se=F, fullrange=F)
#g <- g + geom_abline(data=coefs, aes(intercept=a, slope=b))
g
# reduced model and coefs
fit.red <- lm(LifeExp ~ log(PPGDP) + africa, life)
b0 <- coef(fit.red)
# plot 
df <- data.frame(b0 = c(b0[1], b0[1]+b0[3]), b1 = c(b0[2],b0[2]))
g <- ggplot(data = life, aes(x=log(PPGDP), y=LifeExp))
g <- g + geom_point(aes(shape = Group, colour = Group), size = 4) + theme_bw()
g <- g + geom_abline(data=df, aes(intercept=b0, slope=b1))
g
# plot model diagnostics
par(mfrow=c(1,2))
plot(fit.red, which = c(1,2))
# F test
sse.fm <- anova(fit.all)[[2]][6]
sse.rm <- anova(fit.red)[[2]][3]
mse.rm <- sse.rm/anova(fit.red)[[1]][3]
f.stat <- (sse.rm - sse.fm)/(mse.rm)
F0 <- qf(0.95, 3, 196)
f.p <- 1 - pf(f.stat, 3, 196)
```
