---
title: "Exam 2"
author: "Cody Frisby"
date: "March 16, 2016"
output: pdf_document
---

### 1)  Depth of Cut and Feed Rate  

```{r, echo = FALSE, message=F, warning=F, cache=T}
feed <- c(rep(0.2,12), rep(0.25,12), rep(0.3,12))
depth <- rep(c(0.15,0.18,0.2, 0.25), 3)
y <- c(74,79,82,99,64,68,88,104,60,73,92,96,92,98,99,104,86,104,108,110,88,88,95,99,99,104,108,114,98,99,110,111,102,95,99,107)
D <- data.frame(y, feed, depth)
D$feed <- as.factor(D$feed)
D$depth <- as.factor(D$depth)
library(dplyr)
y.i <- D %>%
  group_by(feed) %>%
  summarise_each(funs(sum), y)
y.j <- D %>% 
  group_by(depth) %>%
  summarise_each(funs(sum), y)
yij. <- D %>% 
  group_by(feed,depth) %>%
  summarise_each(funs(sum), y)
y.. <- sum(y)
a <- 3; b <- 4; n <- 3; N <- a*b*n
ssa <- (1/(b*n))*sum(y.i$y^2) - y..^2/N 
ssb <- (1/(a*n))*sum(y.j$y^2) - y..^2/N
sssub <- (1/n)*sum(yij.$y^2) - y..^2/N
ssab <- sssub - ssa - ssb
sst <- sum(y^2) - y..^2/N
sse <- sst - ssab - ssa - ssb
msa <- ssa/(a-1); msb <- ssb/(b-1); msab <- ssab/((a-1)*(b-1))
mse <- sse/(a*b*(n-1))
Fa <- msa/mse; Fb <- msb/mse; Fab <- msab/mse
df <- a*b*(n-1)
p.Fa <- 1 - pf(Fa, a-1, df); p.Fb <- 1 - pf(Fb, b-1, df)
p.Fab <- 1 - pf(Fab, (a-1)*(b-1), df)
fit.1 <- aov(y ~ feed*depth, D)
#knitr::kable(anova(fit))
```

#### A) Model  
The model for this experiment is a two factor, three & four level factorial experiment where we have 3 replicates.  The statistical model is represented by
$$y_{ijk} = \mu + \tau_i + \beta_j + (\tau\beta)_{ij} + \varepsilon_{ijk}\left\{
        \begin{array}{ll}
            i = 1, 2, 3 \\
            j = 1, 2, 3, 4 \\
            k = 1, 2, 3
        \end{array}
    \right.
    $$
where $\mu$ is the overall mean of the response: **finish**, $\tau_i$ is the *i*th effect from **feed rate**, $\beta_j$ is the *j*th effect from **depth of cut**, $(\tau\beta)_{ij}$ is the effect from the interaction of **feed rate** and **depth of cut**, and $\varepsilon_{ijk}$ are the errors.  Here we assume the errors are independent, identically distributed ~$N(0, \sigma^2)$.

#### B)  ANOVA  
We construct an ANOVA table based on the above model and model assumptions.  

|Source| Sum Sq | Df | Mean Sq |F value|Pr(>F)|
|:------|:------|:-----:|:---------|:------:|------:|
|feed|$`r ssa`$|$`r a-1`$|$`r msa`$|$`r Fa`$|$`r p.Fa`$|
|depth|$`r ssb`$|$`r b-1`$|$`r msb`$|$`r Fb`$|$`r p.Fb`$|
|feed:depth| $`r ssab`$ | $`r (a-1)*(b-1)`$ | $`r msab`$| $`r Fab`$|$`r p.Fab`$ |
|Error|`r sse`| `r a*b*(n-1)`|$`r mse`$| | 
|Total|$`r sst`$  |  `r N-1`|   |  | 

We conclude that both factors have a significant effect on the response as does their interaction.  

#### C) Model Adequacy  

```{r, echo=F, dpi=200, fig.align='center', fig.height=3}
par(mfrow=c(1,2))
plot(fit.1, which = c(1,2))
```

```{r, echo=F, dpi=200, fig.align='center', fig.height=3}
par(mfrow=c(1,2))
plot(fit.1$residuals~as.numeric(D$feed), xlab = "feed rate", 
     ylab = "Residuals"); abline(h=0, lty = 3, col = "red")
plot(fit.1$residuals~as.numeric(D$depth), xlab = "depth of cut", 
     ylab = "Residuals"); abline(h=0, lty = 3, col = "red")
```

There doesn't appear to be any violations of our models assumptions.  Normality of the errors looks OK as well as equality of variances.  

#### D) Point Estimates and Confidence Intervals - Feed Rate  

```{r, echo=F}
est.25 <- mean(D[D$feed == 0.25, 1])
est.3 <- mean(D[D$feed == 0.3, 1])
t0 <- qt(0.975, a*b*(n-1))
cv <- sqrt(mse/n)
cv2 <- sqrt(2*mse/n)
```

The point estimate for when feed is equal to **0.25** is `r est.25`.  We calculate a confidence interval by  $$\hat y \pm t_{\alpha/2:df} \sqrt{\frac{MS_E}{n}}$$
$$`r est.25` \pm `r t0` \sqrt{\frac{`r mse`}{`r n`}}$$  which results in a 95% confidence interval of $[`r est.25 - t0*cv`, `r est.25 + t0*cv`]$ for $feed = 0.25$.  We would do the same for when feed rate is equal to 0.3, `r est.3`, resulting in a 95% confidence interval of $[`r est.3 - cv*t0`, `r est.3 + cv*t0`]$.  

#### E) Point Estimates and Confidence Intervals - Depth of Cut Differences

```{r, echo=F}
estd.2 <- mean(D[D$depth == 0.2, 1])
estd.25 <- mean(D[D$depth == 0.25, 1])
est <- estd.2 - estd.25
```

Similarly to part D, we first estimate the points when **depth of cut** is equal to 0.2 and 0.25, `r estd.2` and `r estd.25` respectively.  We then calculate a confidence interval using the same formula as above but taking the difference between the two point estimates and multiplying $MS_E$ by 2.  $$`r estd.2` - `r estd.25` \pm (`r t0`)\sqrt{\frac{2*`r mse`}{`r n`}}$$  $$[`r est - cv2*t0`, `r est + cv2*t0`]$$

### 2) Density of Particle Board  
#### A)  
There are no replicates, *ie* $n = 1$, so the error degrees of freedom would result in zero since $abc(n-1) = 0$ for any abc. 

```{r, echo=F}
# get the data into R.  There isn't a file so we will type it in
y <- c(40.9,41.9,42,44.4,46.2,48.4,42.8,43.9,44.8,48.2,48.6,50.7,
       45.4,46,46.2,49.9,50.8,50.3)
resin <- as.factor(c(rep(6,6), rep(9,6), rep(12,6)))
slash <- as.factor(rep(c(0,25,50), 6))
target <- as.factor(rep(c(rep(42,3),rep(48,3)), 3))
G <- data.frame(y, resin, slash, target)
```

#### B) Main Effect Target Estimate  

```{r, echo=F}
est.t <- mean(G[G$target == 48,1]) - mean(G[G$target == 42,1])
tar.48 <- G[G$target == 48,1]
tar.42 <- G[G$target == 42,1]
```

The main effect estimate for target is computed by taking the mean of y when target is equal to 48 minus the mean of y when target is equal to 42.  $$\frac{`r sum(tar.48)`}{`r length(tar.48)`} - \frac{`r sum(tar.42)`}{`r length(tar.42)`} = `r est.t`$$

#### C) ANOVA  

*Here, I display the ANOVA table for the model from this problem.  The calculations were done in R by "hand".  I will display the code at the end of this document.*

```{r, echo=F, message=F, warning=F}
library(dplyr)
# using R to calculate the model
fit.G <- aov(y ~ target+resin*slash, data = G)
# build the ANOVA table by "hand"
y..k <- G %>% # blocking
  group_by(target) %>%
  summarise_each(funs(sum), y)
y.i <- G %>% # rows
  group_by(resin) %>%
  summarise_each(funs(sum), y)
y.j <- G %>% # columns
  group_by(slash) %>%
  summarise_each(funs(sum), y)
yij. <- G %>% # subgroups
  group_by(resin, slash) %>%
  summarise_each(funs(sum), y)
y.. <- sum(y) # sum of all obs
a <- 3; b <- 3; n <- 2; N <- a*b*n
blocks <- (1/(a*b))*sum(y..k$y^2) - y..^2/N #ss due to blocks
ssa <- (1/(b*n))*sum(y.i$y^2) - y..^2/N #ss due to row
ssb <- (1/(a*n))*sum(y.j$y^2) - y..^2/N # ss due to col
sssub <- (1/n)*sum(yij.$y^2) - y..^2/N # ss sub for interaction
ssab <- sssub - ssa - ssb # ss due to interaction
sst <- sum(y^2) - y..^2/N # total sum of squares
sse <- sst - ssab - ssa - ssb - blocks # ss errors
msa <- ssa/(a-1); msb <- ssb/(b-1); msab <- ssab/((a-1)*(b-1))
msblocks <- blocks/(n-1)
mse <- sse/((a*b-1)*(n-1))
sigdelta <- (msblocks - mse)/(a*b)
Fa <- msa/mse; Fb <- msb/mse; Fab <- msab/mse; Fbl <- msblocks/mse
df <- (a*b-1)*(n-1)
p.Fa <- 1 - pf(Fa, a-1, df); p.Fb <- 1 - pf(Fb, b-1, df)
p.Fab <- 1 - pf(Fab, (a-1)*(b-1), df)
p.Fbl <- 1 - pf(Fbl, n-1, df)
```


|Source| Sum Sq | Df | Mean Sq |F value|Pr(>F)|
|:------|:------|:-----:|:---------|:------:|------:|
|target|$`r blocks`$|$`r n-1`$|$`r msblocks`$|$`r Fbl`$|$`r p.Fbl`$|
|resin|$`r ssa`$|$`r a-1`$|$`r msa`$|$`r Fa`$|$`r p.Fa`$|
|slash|$`r ssb`$|$`r b-1`$|$`r msb`$|$`r Fb`$|$`r p.Fb`$|
|resin:slash | $`r ssab`$ | $`r (a-1)*(b-1)`$ | $`r msab`$| $`r Fab`$|$`r p.Fab`$ |
|Error|`r sse`| `r a*b*(n-1)`|$`r mse`$| | 
|Total|$`r sst`$  |  `r N-1`|   |  | 

Testing our model, at $\alpha = 0.01$, we conclude that all the effects are significant *except* the interaction of resin & slash.

#### D) "Sweet Spot"  

The best treatment combinations for minimizing the response would be resin = 6, slash = 0, target = 42.  

```{r, echo=F, dpi=200, fig.align='center'}
G$pred <- fit.G$fitted.values 
# append the fitted values to our data table.  Then we can look at 
# what the factor levels are when we predict y at its lowest.  There
# are ways to plot this...it's a little more complex then I want to
# spend time on at the moment.
```

### 3)  Experiment with three factors with two levels and 4 replicates.

A table listing the main effects of a, b, and c respectively.

```{r, echo=F}
one <- 425; a <- 426; b <- 1118; ab <- 1203; c <- 1283 
ac <- 1396; bc <- 1670; abc <- 1807; n <- 4
# Part A, main effects
main.a <- (1/(4*n))*(a-one+ab-b+ac-c+abc-bc)
main.b <- (1/(4*n))*(b+ab+bc+abc-one-a-c-ac)
main.c <- (1/(4*n))*(c+ac+bc+abc-one-a-b-ab)
# table of the main effects
knitr::kable(cbind(main.a, main.b, main.c))
```

#### B) AB interaction effect  

```{r, echo=F}
k <- 3; n <- 4
AB <- (1/(4*n))*(abc-bc+ab-b-ac+c-a+one)
ssab <- (abc-bc+ab-b-ac+c-a+one)^2 / 32
ssa <- (a-one+ab-b+ac-c+abc-bc)^2 / 32
ssb <- (b+ab+bc+abc-one-a-c-ac)^2 / 32
ssc <- (c+ac+bc+abc-one-a-b-ab)^2 / 32
ssbc <- (one+a-b-ab-c-ac+bc+abc)^2 / 32
ssac <- (one-a+b-ab-c+ac-bc+abc)^2 / 32
ssabc <- (abc-bc-ac+c-ab+b+a-one)^2 / 32
sse <- 466150-ssa-ssb-ssc-ssab-ssac-ssbc-ssabc
dfe <- 2^k*(n-1); dft <- 2^k*n - 1
mse <- sse/dfe; Fa <- ssa/mse; Fb <- ssb/mse; Fc <- ssc/mse
Fab <- ssab/mse; Fac <- ssac/mse; Fbc <- ssbc/mse; Fabc <- ssabc/mse
```

The estimate for the AB interaction is `r AB`.  

#### C) ANOVA Table  

Here, using a computer, we fill in the ANOVA table.  

|Source| Sum Sq | Df | Mean Sq |F value|
|:------|:------|:-----:|:--------|:------:|
|A|$`r ssa`$|$1$|$`r ssa`$|$`r Fa`$|
|B|$`r ssb`$|$1$|$`r ssb`$|$`r Fb`$|
|C|$`r ssc`$|$1$|$`r ssc`$|$`r Fc`$|
|AB| $`r ssab`$|$1$|$`r ssab`$| $`r Fab`$|
|AC|$`r ssac`$|$1$|$`r ssac`$|$`r Fac`$|
|BC|$`r ssab`$|$1$|$`r ssab`$|$`r Fab`$|
|ABC|$`r ssabc`$|$1$|$`r ssabc`$|$`r Fabc`$|
|Error|`r sse`| `r dfe`|$`r mse`$| | 
|Total|$466150$|$`r dft`$|   |  | 

Critical F value for $\alpha = 0.01$ is $`r qf(0.99,1,dfe)`$, so for any F value greater than this we would conclude that effect is significant.  From the above ANOVA table it appears we have three, they are the main effects A, B, and C.  

#### D)  Factor Conclusions  

It appears the main effects have a significant effect on the response while their interaction with each other does not.  B and C appear to have a stronger affect than A. 

#### E)  Best for Maximizing yield  

I would say a+ b+ c+ since the main effects of the three treatments are all positive.  

### 4)  Marketing print experiment  

```{r, echo=F}
y <- c(50,54,44,42,46,48,42,43,49,46,48,45,47,48,56,54)
A <- as.factor(rep(c(rep(3,2), rep(1,2)), 4))
B <- as.factor(rep(c(rep("BW", 4), rep("Color",4)), 2))
C <- as.factor(rep(c(rep(19.95,8), rep(24.95,8)), 1))
H <- data.frame(y, A, B, C)
fit.H <- aov(y ~ A*B*C, data = H)
```

#### A) ANOVA  
Here, we look at an ANOVA table based on the model of main effects and interaction effects between all three variables.  Here, $n = 2$, we have 2 replicates and $2^k$, where $k = 3$, factor/level combinations.  

`r knitr::kable(anova(fit.H))`   

It appears that the most significant factor is the *interaction* between **price** and **class** of the direct mail.  Also significant are main effect C (price), AB interaction, and BC interaction.  

#### B) Model Adequacy  

Lets take a look at the model residuals.  

```{r, echo=F, dpi=200, fig.align='center', fig.height=3}
par(mfrow=c(1,2))
plot(fit.H, which = c(1,2))
```

```{r, echo=F, dpi=200, fig.align='center', fig.height=3}
par(mfrow=c(1,2))
plot(fit.H$residuals~as.numeric(H$A), xlab = "Class of Mail", 
     ylab = "Residuals"); abline(h=0, lty = 3, col = "red")
plot(fit.H$residuals~as.numeric(H$B), xlab = "Print Color", 
     ylab = "Residuals"); abline(h=0, lty = 3, col = "red")
plot(fit.H$residuals~as.numeric(H$C), xlab = "Print Price", 
     ylab = "Residuals"); abline(h=0, lty = 3, col = "red")
```

Model assumptions are OK.  There's nothing that need concern us from the residual plots.  

#### C)  Recomendations  

If we wish to maximize orders then the reccomendation would be to print in color and send first class.  If there is a desire to have a lower price ($19.95) while still maximizing orders, then the reccomendation would be to print in black and white and send 3rd class.  



### R Code:

```{r, eval=F}
feed <- c(rep(0.2,12), rep(0.25,12), rep(0.3,12))
depth <- rep(c(0.15,0.18,0.2, 0.25), 3)
y <- c(74,79,82,99,64,68,88,104,60,73,92,96,92,98,99,104,86,104,108,110,88,88,95,99,99,104,108,114,98,99,110,111,102,95,99,107)
D <- data.frame(y, feed, depth)
D$feed <- as.factor(D$feed)
D$depth <- as.factor(D$depth)
library(dplyr)
y.i <- D %>%
  group_by(feed) %>%
  summarise_each(funs(sum), y)
y.j <- D %>% 
  group_by(depth) %>%
  summarise_each(funs(sum), y)
yij. <- D %>% 
  group_by(feed,depth) %>%
  summarise_each(funs(sum), y)
y.. <- sum(y)
a <- 3; b <- 4; n <- 3; N <- a*b*n
ssa <- (1/(b*n))*sum(y.i$y^2) - y..^2/N 
ssb <- (1/(a*n))*sum(y.j$y^2) - y..^2/N
sssub <- (1/n)*sum(yij.$y^2) - y..^2/N
ssab <- sssub - ssa - ssb
sst <- sum(y^2) - y..^2/N
sse <- sst - ssab - ssa - ssb
msa <- ssa/(a-1); msb <- ssb/(b-1); msab <- ssab/((a-1)*(b-1))
mse <- sse/(a*b*(n-1))
Fa <- msa/mse; Fb <- msb/mse; Fab <- msab/mse
df <- a*b*(n-1)
p.Fa <- 1 - pf(Fa, a-1, df); p.Fb <- 1 - pf(Fb, b-1, df)
p.Fab <- 1 - pf(Fab, (a-1)*(b-1), df)
fit.1 <- aov(y ~ feed*depth, D)
#knitr::kable(anova(fit))
par(mfrow=c(1,2))
plot(fit.1, which = c(1,2))
par(mfrow=c(1,2))
plot(fit.1$residuals~as.numeric(D$feed), xlab = "feed rate", 
     ylab = "Residuals"); abline(h=0, lty = 3, col = "red")
plot(fit.1$residuals~as.numeric(D$depth), xlab = "depth of cut", 
     ylab = "Residuals"); abline(h=0, lty = 3, col = "red")
est.25 <- mean(D[D$feed == 0.25, 1])
est.3 <- mean(D[D$feed == 0.3, 1])
t0 <- qt(0.975, a*b*(n-1))
cv <- sqrt(mse/n)
cv2 <- sqrt(2*mse/n)
estd.2 <- mean(D[D$depth == 0.2, 1])
estd.25 <- mean(D[D$depth == 0.25, 1])
est <- estd.2 - estd.25
# get the data into R.  There isn't a file so we will type it in
y <- c(40.9,41.9,42,44.4,46.2,48.4,42.8,43.9,44.8,48.2,48.6,50.7,
       45.4,46,46.2,49.9,50.8,50.3)
resin <- as.factor(c(rep(6,6), rep(9,6), rep(12,6)))
slash <- as.factor(rep(c(0,25,50), 6))
target <- as.factor(rep(c(rep(42,3),rep(48,3)), 3))
G <- data.frame(y, resin, slash, target)
est.t <- mean(G[G$target == 48,1]) - mean(G[G$target == 42,1])
tar.48 <- G[G$target == 48,1]
tar.42 <- G[G$target == 42,1]
library(dplyr)
# using R to calculate the model
fit.G <- aov(y ~ target+resin*slash, data = G)
# build the ANOVA table by "hand"
y..k <- G %>% # blocking
  group_by(target) %>%
  summarise_each(funs(sum), y)
y.i <- G %>% # rows
  group_by(resin) %>%
  summarise_each(funs(sum), y)
y.j <- G %>% # columns
  group_by(slash) %>%
  summarise_each(funs(sum), y)
yij. <- G %>% # subgroups
  group_by(resin, slash) %>%
  summarise_each(funs(sum), y)
y.. <- sum(y) # sum of all obs
a <- 3; b <- 3; n <- 2; N <- a*b*n
blocks <- (1/(a*b))*sum(y..k$y^2) - y..^2/N #ss due to blocks
ssa <- (1/(b*n))*sum(y.i$y^2) - y..^2/N #ss due to row
ssb <- (1/(a*n))*sum(y.j$y^2) - y..^2/N # ss due to col
sssub <- (1/n)*sum(yij.$y^2) - y..^2/N # ss sub for interaction
ssab <- sssub - ssa - ssb # ss due to interaction
sst <- sum(y^2) - y..^2/N # total sum of squares
sse <- sst - ssab - ssa - ssb - blocks # ss errors
msa <- ssa/(a-1); msb <- ssb/(b-1); msab <- ssab/((a-1)*(b-1))
msblocks <- blocks/(n-1)
mse <- sse/((a*b-1)*(n-1))
sigdelta <- (msblocks - mse)/(a*b)
Fa <- msa/mse; Fb <- msb/mse; Fab <- msab/mse; Fbl <- msblocks/mse
df <- (a*b-1)*(n-1)
p.Fa <- 1 - pf(Fa, a-1, df); p.Fb <- 1 - pf(Fb, b-1, df)
p.Fab <- 1 - pf(Fab, (a-1)*(b-1), df)
p.Fbl <- 1 - pf(Fbl, n-1, df)
G$pred <- fit.G$fitted.values 
one <- 425; a <- 426; b <- 1118; ab <- 1203; c <- 1283 
ac <- 1396; bc <- 1670; abc <- 1807; n <- 4
# Part A, main effects
main.a <- (1/(4*n))*(a-one+ab-b+ac-c+abc-bc)
main.b <- (1/(4*n))*(b+ab+bc+abc-one-a-c-ac)
main.c <- (1/(4*n))*(c+ac+bc+abc-one-a-b-ab)
# table of the main effects
knitr::kable(cbind(main.a, main.b, main.c))
k <- 3; n <- 4
AB <- (1/(4*n))*(abc-bc+ab-b-ac+c-a+one)
ssab <- (abc-bc+ab-b-ac+c-a+one)^2 / 32
ssa <- (a-one+ab-b+ac-c+abc-bc)^2 / 32
ssb <- (b+ab+bc+abc-one-a-c-ac)^2 / 32
ssc <- (c+ac+bc+abc-one-a-b-ab)^2 / 32
ssbc <- (one+a-b-ab-c-ac+bc+abc)^2 / 32
ssac <- (one-a+b-ab-c+ac-bc+abc)^2 / 32
ssabc <- (abc-bc-ac+c-ab+b+a-one)^2 / 32
sse <- 466150-ssa-ssb-ssc-ssab-ssac-ssbc-ssabc
dfe <- 2^k*(n-1); dft <- 2^k*n - 1
mse <- sse/dfe; Fa <- ssa/mse; Fb <- ssb/mse; Fc <- ssc/mse
Fab <- ssab/mse; Fac <- ssac/mse; Fbc <- ssbc/mse; Fabc <- ssabc/mse
y <- c(50,54,44,42,46,48,42,43,49,46,48,45,47,48,56,54)
A <- as.factor(rep(c(rep(3,2), rep(1,2)), 4))
B <- as.factor(rep(c(rep("BW", 4), rep("Color",4)), 2))
C <- as.factor(rep(c(rep(19.95,8), rep(24.95,8)), 1))
H <- data.frame(y, A, B, C)
fit.H <- aov(y ~ A*B*C, data = H)
par(mfrow=c(1,2))
plot(fit.H, which = c(1,2))
par(mfrow=c(1,2))
plot(fit.H$residuals~as.numeric(H$A), xlab = "Class of Mail", 
     ylab = "Residuals"); abline(h=0, lty = 3, col = "red")
plot(fit.H$residuals~as.numeric(H$B), xlab = "Print Color", 
     ylab = "Residuals"); abline(h=0, lty = 3, col = "red")
plot(fit.H$residuals~as.numeric(H$C), xlab = "Print Price", 
     ylab = "Residuals"); abline(h=0, lty = 3, col = "red")
```
