---
title: "Homework 9"
author: "Cody Frisby"
date: "April 11, 2016"
output: pdf_document
---

```{r, echo=F, message=F, warning=F}
####### Exploring 3^2 full factorial models using R ########
df <- read.csv("~/Documents/STAT4100/data/hw9.csv")
df.c <- df
fit.lm <- lm(y ~ glass*temp, df) # fit with continuous vars
fit.aov.c <- aov(y ~ .*., df)
df$glass <- as.factor(df$glass)
df$temp <- as.factor(df$temp)
fit.aov <- aov(y ~ .*., df) # need factor variables
#summary(fit.aov)
#summary(fit.aov.c)
# R package for response surface models
library(rsm)
fit.rsm <- rsm(y ~ SO(glass,temp), df.c) 
# predictors need to be continuous.
# also, the anova(fit.rsm) does not seperate
# out the individual SS terms.....lame
# to use R and see the SS terms for each term in the model...
# ... first create new variables
df.c$glassglass <- df.c$glass^2
df.c$temptemp <- df.c$temp^2
fit.full <- aov(y ~ .*., df.c)
# the below 2 values should be equal
#sum(anova(fit.full)[1:2,2]);anova(fit.rsm)[1,2]
# yes! they are, what about...
#anova(fit.rsm)[3,2]; sum(anova(fit.full)[c(3,4),2])
# Yes!, so what about the other interaction terms...
# I believe fit.full has too many terms...I may need to 
# fix this
fit.full2 <- aov(y ~ glass*temp+temptemp+glassglass, df.c)
# this model results in the same value for residuals as the 
# fit.rsm model....may be on the right track now.
# basically the same models...minus rounding error
#anova(fit.full2, fit.rsm)
# tool life example from textbook
tool <- read.table("~/Documents/STAT4100/data/toollife.txt", 
                   header = TRUE)
tool$Angle2 <- tool$Angle^2; tool$Speed2 <- tool$Speed^2

#anova(aov(Life ~ Angle*Speed, tool))
#anova(aov(Life ~ Angle*Speed+Speed2+Angle2, tool))
# or like 9.1.2
#anova(aov(Life ~ .*., tool))
```

```{r, echo=F, warning=F, message=F}
# this is really long, I really need to write a function
# one interaction term, no squared terms
library(dplyr)
y <- df$y
y.i <- df.c %>%
  group_by(glass) %>%
  summarise_each(funs(sum), y)
y.j <- df.c %>%
  group_by(temp) %>%
  summarise_each(funs(sum), y)
yij. <- df.c %>%
  group_by(glass, temp) %>%
  summarise_each(funs(sum), y)
y.. <- sum(y)
a <- 3; b <- 3; n <- 3; N <- a*b*n
ssa <- (1/(b*n))*sum(y.i$y^2) - y..^2/N 
ssb <- (1/(a*n))*sum(y.j$y^2) - y..^2/N
sssub <- (1/n)*sum(yij.$y^2) - y..^2/N
ssab <- sssub - ssa - ssb
sst <- sum(y^2) - y..^2/N
sse <- sst - ssab - ssa - ssb
#round(c(ssa,ssb,ssab,sse,sst))
# now to add squared terms
row.sums <- df %>%
  group_by(glass,temp) %>%
  summarise_each(funs(sum), y)
y <- row.sums$y
id <- c("00", "01", "02", "10", "11", "12", "20", "21", "22")
#data.frame(id, y) # these are our summaries..
#... now to develop the matrix
Al <- c(rep(-1,3), rep(0,3), rep(1,3))
Bl <- rep(c(-1,0,1),3)
Aq <- c(rep(1,3), rep(-2,3), rep(1,3))
Bq <- rep(c(1,-2,1), 3)
AlBl <- Al*Bl; AlBq <- Al*Bq; AqBl <- Aq*Bl; AqBq <- Aq*Bq
#cbind(Al,Bl,Aq,Bq,AlBl,AlBq,AqBl,AqBq) # our matrix
dm <- cbind(Al,Bl,Aq,Bq,AlBl,AlBq,AqBl,AqBq)
rownames(dm) <- id
m.effects <- apply(dm,2, FUN = function(d) sum(d*y)/3) # effects
# now for the sum of squares...
C <- apply(dm, 2, function(x) sum(x^2)) #contrasts
ss <- 3*(m.effects^2)/C # sum of squares
ss2 <- round(cbind(ssa,ssb,ssab,sse,sst))
sse <- sst - sum(ss); mse <- sse/18
```

Here, I did all the Calculations for the ANOVA table by "hand".  **Note**: *all R code will be at the end of this document*.

First, I display the contrasts matrix `r knitr::kable(dm)` and with the sum of each treatment levels' combinations `r knitr::kable(cbind(dm,y))`  Now, we can estimate the effect of each term in our model by multiplying *y* by each column of this matrix, taking the sum, and dividing by $n = 3$.  The result is `r knitr::kable(m.effects)`  Now, for the Sums of Squares for each term in the model we calculate $$\frac{n(contrasts)^2}{\sum{c_i^2}}$$ which results in `r knitr::kable(ss)` Before I forget, the sum of squares for A, B, and AB is `r knitr::kable(ss2)` The remaining ANOVA table is similar to previous methods used in this class, where mean squares are $\frac{SS}{df}$ and the F statistic is $\frac{MS_i}{MS_E}$.  Here, I finish filling out the ANOVA table   

|Source| Sum Sq | Df | Mean Sq |F value|Pr(>F)|
|:------|:------|:-----:|:---------|:------:|------:|
|glass|$`r ss[1]`$| 1 |$`r ss[1]`$|$`r ss[1]/mse`$|<0.000|
|temp|$`r ss[2]`$| 1 |$`r ss[2]`$|$`r ss[2]/mse`$|<0.000|
|$glass^2$|$`r ss[3]`$ | 1 | $`r ss[3]`$|$`r ss[3]/mse`$|<0.000|
|$temp^2$|$`r ss[4]`$| 1 |$`r ss[4]`$|$`r ss[4]/mse`$|<0.000|
|temp:glass|$`r ss[5]`$| 1 |$`r ss[5]`$|$`r ss[5]/mse`$| <0.000|
|$glass:temp^2$|$`r ss[6]`$ | 1 |$`r ss[6]`$|$`r ss[6]/mse`$|<0.000| 
|$glass^2:temp$|$`r ss[7]`$| 1 |$`r ss[7]`$|$`r ss[7]/mse`$|<0.000|
|$glass^2:temp^2$|$`r ss[8]`$| 1 |$`r ss[8]`$|$`r ss[8]/mse`$|<0.000|
|Error|$`r sse`$| 18 | $`r mse`$ | | |
|Total|$`r sst`$| 26 | | | |

All effects have a significant effect on *light output* and are statistically significant, with **glass** having the most significant effect.  Theere is also some *curvature* to the main effects **glass** and **temp**

  Now, I take a look at the residual plots of the model

```{r, dpi=200, fig.align='center', warning=F, message=F, echo=F}
par(mfrow=c(1,2))
plot(fit.full, which = c(1,2))
```

The normality assumption may not be valid.  We may need to reexamine the model or possibly perform a transformation.  

### R code:  

```{r, eval=F, message=F, warning=F}
####### Exploring 3^2 full factorial models using R ########
df <- read.csv("~/Documents/STAT4100/data/hw9.csv")
df.c <- df
library(dplyr)
y <- df$y
y.i <- df.c %>%
  group_by(glass) %>%
  summarise_each(funs(sum), y)
y.j <- df.c %>%
  group_by(temp) %>%
  summarise_each(funs(sum), y)
yij. <- df.c %>%
  group_by(glass, temp) %>%
  summarise_each(funs(sum), y)
y.. <- sum(y)
a <- 3; b <- 3; n <- 3; N <- a*b*n
ssa <- (1/(b*n))*sum(y.i$y^2) - y..^2/N 
ssb <- (1/(a*n))*sum(y.j$y^2) - y..^2/N
sssub <- (1/n)*sum(yij.$y^2) - y..^2/N
ssab <- sssub - ssa - ssb
sst <- sum(y^2) - y..^2/N
sse <- sst - ssab - ssa - ssb
#round(c(ssa,ssb,ssab,sse,sst))
# now to add squared terms
row.sums <- df %>%
  group_by(glass,temp) %>%
  summarise_each(funs(sum), y)
y <- row.sums$y
id <- c("00", "01", "02", "10", "11", "12", "20", "21", "22")
#data.frame(id, y) # these are our summaries..
#... now to develop the contrasts matrix
Al <- c(rep(-1,3), rep(0,3), rep(1,3))
Bl <- rep(c(-1,0,1),3)
Aq <- c(rep(1,3), rep(-2,3), rep(1,3))
Bq <- rep(c(1,-2,1), 3)
AlBl <- Al*Bl; AlBq <- Al*Bq; AqBl <- Aq*Bl; AqBq <- Aq*Bq
dm <- cbind(Al,Bl,Aq,Bq,AlBl,AlBq,AqBl,AqBq)
rownames(dm) <- id
m.effects <- apply(dm,2, FUN = function(d) sum(d*y)/3) # effects
# now for the sum of squares...
C <- apply(dm, 2, function(x) sum(x^2)) #contrasts
ss <- 3*(m.effects^2)/C # sum of squares
ss2 <- round(cbind(ssa,ssb,ssab,sse,sst))
sse <- sst - sum(ss); mse <- sse/18
par(mfrow=c(1,2))
plot(fit.full, which = c(1,2))
```