---
title: "Sin(x) = x Fixed Point Iteration"
author: "Cody Frisby"
date: "9/11/2017"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = "")
```

First I define a function called `fixed_plot` where I combine fixed point iteration and plotting (I'm a visual guy).  

```{r define function, echo=FALSE}
fixed_plot <- function(x, FUN = cos, n = 10) {
  xmin <- -abs(x)
  xmax <- abs(x)
  plot(1:abs(x), xlim = c(xmin - 1, xmax + 1), 
       ylim = c(-2, 2), type = "n", xlab = "", ylab = "")
  grid()
  abline(h = 0, lty = 3) # x axis
  abline(v = 0, lty = 3)# y axis
  abline(coef = c(0, 1), col = "dodgerblue", lty = 3)
  curve(FUN(x), from = xmin - 2, xmax + 2, lwd = 2, add = TRUE)
  x0 <- x
  i <- 1
  m <- matrix(ncol = 2)
  for(i in 1:n) {
    x0 <- FUN(x) # calculate f(x)
    r <- x0/x
    #a <- 1 - (x^2/6)
    points(x, x0, col = i)
    x <- x0
    #print(x); print(r); print(a)
    #Sys.sleep(0.1)
    m <- rbind(m, c(x0, r))
  }
  points(x0, FUN(x0), pch = "X", cex = 2)
  m <- m[-1, ] # remove the stupid NA row
  colnames(m) <- c("x0", "ratio errors")
  return(m)
}
```


Notice that I have multiplied $sin(x)$ by $\frac{x^3}{6}$
which makes $$f(x) = sin(x) \frac{x^3}{6}$$ 

which you can easily identify as the second term in the taylor series expansion of $sin(x)$.  

So, this little "fudge function" makes my function converge super fast now, but ONLY if my starting point is less than $|3.6|$.  If I choose any number greater than that my iteration diverges, and rapidly.  Here's some examples with only 5 iterations.  

```{r}
fixed_plot(3.6178, FUN = sin, n = 10)
```

  I'm at **zero** super fast as you can see.  Now try a starting guess greater than that.  


```{r}
fixed_plot(3.61781, FUN = sin, n = 10)
```


**And we diverge super fast**.  


What's going on here?  
