---
title: "Lab Notes"
author: "Cody"
date: "8/24/2017"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.align = 'center', dpi = 200)
```


Rate of change in position over the rate of change in time is defined as velocity.  

$$\frac{dx}{dt} = v$$

$$\frac{d^2x}{dt^2} = \frac{dv}{dt} = a$$

$$\frac{x_{n+1} - x_n}{\delta t} = v$$

$$x_{n+1} = v_0 \delta t + x_n$$  

$$a = \frac{v_{n+1} - v_n}{\delta t}$$  

Euler's method.  Verlay's method.  

Which do you update first?  Position or velocity?


## Lab0  

First, I'm going to reproduct the time and $sin(omega t)$ data and plot it.  

```{r}
omega <- 5
s <- 0.025
time <- seq(0, 10, s)
f <- function (t, omega = 3) {
  sin(omega * t)
}
plot(time, f(t = time), type = "l", ylab = "sin(omega * t)")
```

Now, we need to take the numerical derivative of $sin(omega t)$ with respect to time.  

```{r}
# in the lab we used excel and created a time diff by lag 2.  
#dx <- numericDeriv(f(t = time), "time", dir = )
# the above code is wrong
dx <- diff(f(time)) / s / omega
fdx <- diffinv(dx) * s + c(0, f(time[-1]))
plot(time[-1], dx, type = "l", xlab = "time", ylab = ("dx"))
plot(time, fdx, type = "l", xlab = "time", ylab = ("fdx"))
```


Using ggplot2 to put them on the same plot. 

```{r}
# first take the integral of the derivative:
library(ggplot2)
df <- data.frame(cbind(time, f(time), dx, fdx))
names(df) <- c("time", "x", "dx", "fdx")
# xlab = expression(hat(mu)[0]), ylab = expression(alpha^beta),
#  main = expression(paste("Plot of ", alpha^beta, " versus ", hat(mu)[0])))
library(reshape2)
df_melt <- melt(df[-1])
df_melt$time <- df$time
gg <- ggplot(df_melt, aes(x = time, y = value, color = variable))
gg + geom_line() + theme_bw()
```


## 8-31-17  

Lecture is to derive kinematic equations for constant acceleration.  

$$a = \frac{dv}{dt}$$  

$$adt = dv$$  

$$\int adt = \int v$$

$$at + v_0 = v$$  

Moving onto the next one.  We can rewrite v as the time integral of position.  

$$at + v_0 = \frac{dx}{dt}$$  

$$\frac{1}{2}at^2 + v_0t + x_0 = x$$  

The elastic force for a spring is $-k \Delta x$.  

Sinusoidal position is $Asin(\omega t)$.  $v = \omega Acos(\omega t)$ and $a = \omega^2 A sin(\omega t)$.  

7-10 should be our spring constant if we do it right.  


In excel, where A is time and B is pos make it $$\frac{B_{n+2} - B_n}{A_{n+2} - A_n}$$  

## 9-7-17  


## 9-14-17  

Today's lab is all about statistics and error and prediction.  

Need to solve a system of linear equations first.  It's easier in R since I already know how to.  

```{r}
x <- matrix(c(1, 1, 0.510204, -0.051020408), ncol = 2)
y <- c(0.3661, 0.5)
solve(x, y)
pracma::rref(cbind(x, y))
x <- matrix(c(1, 1, (0.510204 + 0.0806), -(0.1)) ,ncol = 2)
y <- c(0.42395, 0.98)
pracma::rref(cbind(x, y))
```

