---
title: "Logistic Regression"
author: "Cody Frisby"
date: "April 2, 2016"
output:
  html_document:
    keep_md: yes
  pdf_document: default
---

###DONNER PARTY  
In 1846, the Donner party (Donner and Reed families) left Springfield, Illinois for California in covered wagons.  After reaching Fort Bridger, Wyoming, the leaders decided to find a new route to Sacramento.  They became stranded in the eastern Sierra Nevada mountains at a place now called Donner Pass when the region was hit by heavy snows in late October.  By the time the survivors were rescued on April 21, 1847, 40 out of the 87 had died.  We will analyze a data set containing the 45 adults (age 18 and over) for this party.

|Variable|Description|
|:------|:------|
|age|Age of individual|
|sex|	1 – male, 0 - female|
|survive|	1 – survived, 0 - died|

####RQ1:  What is the relationship between survival and gender?  
####RQ2:  Predict the probability of survival as a function of age.  
####RQ3:  After taking into account age, are women more likely to survive harsh conditions than men?  

  First, we get the data into R and take a look at some summarys.  

```{r}
don <- read.table("~/Documents/MATH3710/donnerparty/donner.txt")
names(don) <- c("age", "sex", "survive")
summary(don) # look at a summary of the data frame
attach(don) # this "attaches" the data frame to our session
table(sex, survive) # table of sex vs. survive
tmp <- chisq.test(sex, survive, correct = F) 
# chi square test for the same
tmp$expected
tmp
```

##Start Logistic Regression  

The logistic regression function is $$\pi(x) = P[Y = 1 | X = x] = \frac{e^{\beta_0 + \beta_1x}}{1 + e^{\beta_0 + \beta_1x}}$$  and is NOT linear.  
 Here we fit the donner data to a logistic model using R.

```{r}
fit <- glm(survive ~ age, family = binomial(), data = don)
summary(fit)$coef # betas
# or
#anova(fit, test = "Chisq")
# use our model to predict
age <- c(10,20,30,40,50,60)
# odds ratio
exp(predict(fit, data.frame(age)))
odds <- exp(predict(fit, data.frame(age)))
p <- odds/(1+odds)
# odds go down with age. 
knitr::kable(cbind(age,odds,p))
```

Look at a plot of the predicted values vs. age  

```{r, dpi=200, fig.align='center'}
plot(don$age, fit$fitted.values, ylab = "Predicted", xlab = "age",
     pch=16, ylim = c(0,1))
```

And to see the odds Ratio:

```{r}
exp(summary(fit)$coef) # odds ratio
exp(confint(fit)) #confidence interval
```

So the logit model can be written as $$ln(\frac{\pi}{1 - \pi}) = \beta_0 + \beta_1x_1 + ... + \beta_kx_k$$

Notes:  
1. The logarithm of the odds ratio is called a logit.  
2. $0 \leq \pi \leq 1$ but $- \infty < ln(\frac{\pi}{1 - \pi}) < \infty$ which makes the logit more appropriate for an unbounded linear function.  
We use MLE (maximum liklihood estimation) to estimate the values for $\beta_i$.  

## Playing around with the class notes and R functions. 

This is the logistic regression function shared in class and in the notes.  
```{r}
# here we define a function 
logit <- function(x){
  -2*log(2) + log(2)*x
}
# here we use the function above on some inputs
gpa <- c(0,1,2,3,4)
logit(gpa)
# and we raise this here to get the odds
exp(logit(gpa))
odds <- exp(logit(gpa))
# and to go back to probabilities
odds/(1+odds)
p <- odds/(1+odds)
knitr::kable(cbind(gpa,odds,p))
```


Here's how you would do a table of the values using R.  Remeber, we have already attached the data frame donner.  

```{r}
with(don, table(age, survive))
with(don, chisq.test(age, survive))
```


```{r}
fit2 <- glm(survive ~ age+sex, family = binomial(link = "logit"), 
            data = don)
summary(fit2)
```

### Performing the Hosmer -  Lemeshow Goodness of Fit Test  

Install the package "ResourceSelection" by typing in install.packages("ResourceSelection").  

```{r}
library(ResourceSelection)
hl <- hoslem.test(fit$y, fit$fitted.values, g = 8)
hl
# we can also look at some tables
cbind(hl$observed, hl$expected)
```



## Concordance  
thanks to 
http://shashiasrblog.blogspot.com/2014/01/binary-logistic-regression-on-r.html
for this function

```{r}
###########################################################
# Function OptimisedConc : for concordance, discordance, ties
# The function returns Concordance, discordance, and ties
# by taking a glm binomial model result as input.
# Although it still uses two-for loops, it optimises the code
# by creating initial zero matrices
###########################################################
OptimisedConc=function(model)
{
  Data = cbind(model$y, model$fitted.values) 
  ones = Data[Data[,1] == 1,]
  zeros = Data[Data[,1] == 0,]
  conc=matrix(0, dim(zeros)[1], dim(ones)[1])
  disc=matrix(0, dim(zeros)[1], dim(ones)[1])
  ties=matrix(0, dim(zeros)[1], dim(ones)[1])
  for (j in 1:dim(zeros)[1])
  {
    for (i in 1:dim(ones)[1])
    {
      if (ones[i,2]>zeros[j,2])
      {conc[j,i]=1}
      else if (ones[i,2]<zeros[j,2])
      {disc[j,i]=1}
      else if (ones[i,2]==zeros[j,2])
      {ties[j,i]=1}
    }
  }
  Pairs=dim(zeros)[1]*dim(ones)[1]
  PercentConcordance=(sum(conc)/Pairs)*100
  PercentDiscordance=(sum(disc)/Pairs)*100
  PercentTied=(sum(ties)/Pairs)*100
  return(list("Percent Concordance"=PercentConcordance,"Percent Discordance"=PercentDiscordance,"Percent Tied"=PercentTied,"Pairs"=Pairs))
}
```

This function, OptimisedConc, takes one input, your logistic model and returns a list of values. 

```{r}
OptimisedConc(fit2)
```

## ROC curves in R  

Let's install another package for this job, pROC, by calling install.packages("pROC"). 

```{r, fig.align='center', dpi=200}
library(pROC)
roc(fit2$y, fit2$fitted.values)
plot(roc(fit2$y, fit2$fitted.values))
# this is why I love R
```


## Classification Table  

We can create our own classification table using R and a logical statement for the value of pi we'd like to classify on, for example $\pi > 0.5$.  

```{r}
theprobs <- fit$fitted.values
table(fit$y, 1*(theprobs > .5))
```


- Plot of predicted vs age by sex.

```{r, dpi=200, fig.align='center', warning=F, message=F}
# adding a new column to sore
don$gender <- 0
don$predicted <- fit2$fitted.values
don[don$sex == 1, "gender"] <- "male"
don[don$sex == 0, "gender"] <- "female"
don$gender <- factor(don$gender)
library(ggplot2) # using ggplot2 to create this plot
g <- ggplot(data = don, aes(x = age, y = predicted))
g <- g + geom_point(aes(shape = gender, colour = gender), size = 4) + theme_bw()
g <- g + geom_smooth(aes(group=gender, linetype = gender, color = gender), 
                     method = "glm", 
          method.args = list(family = "binomial"), se = F) 
g
```

## Interaction Term, Our third model for Donner Party Data

```{r}
fit3 <- glm(survive ~ age*sex, family = binomial(link = "logit"), 
            data = don)
summary(fit3)
# let's predict for females
age <- c(20,30,40,50,60)
# odds ratio
exp(predict(fit3, data.frame(age, sex = (rep(0,5)))))
odds <- exp(predict(fit3, data.frame(age, sex = rep(0,5))))
p <- odds/(1+odds)
p
# and now for males
exp(predict(fit3, data.frame(age, sex = (rep(1,5)))))
odds <- exp(predict(fit3, data.frame(age, sex = rep(1,5))))
p <- odds/(1+odds)
p
# do we want to get this interaction term in??? Yes!
anova(fit2, fit3, test = "LRT")
```
