---
title: "Chapter 7 Homework"
author: "Cody Frisby"
date: "10/26/2016"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, fig.align = 'center', dpi=200, 
                      warning=F, message = F)
```

## 7.1  

First we plot the P chart without excluding any data.  

```{r}
library(qcc)
old <- qcc.options()
qcc.options(bg.margin = "white")
# 7.1
x <- c(7,4,1,3,6,8,10,5,2,7,6,15,0,9,5,1,4,5,7,12)
# plotting the control limits without removing
pchart <- qcc(x, type = "p", size = 100)


```

Now we plot the control chart excluding any outlier.  

```{r}
# removing the outlier programtically...
x <- x[-pchart$violations[1]$beyond.limits]
pchart2 <- qcc(x, type="p", size = 100)
```


## 7.2  

```{r}
x <- c(8,1,3,0,2,2,4,0,1,10,6,6,0,4,0,3,1,15,2,3,0)
pchart <- qcc(x, type = "p", size = 150)
```

It appears the process is not operating in control.  In our samples, there are two samples that are out of control, beyond the UCL, as can be seen from the control chart.  

Removing the points that are OOC we get
```{r}
x <- x[-pchart$violations[1]$beyond.limits]
pchart2 <- qcc(x, type="p", size = 150)
```

and our new reccomended UCL and LCL are displayed on the chart.  

## 7.6  

```{r}
x <- c(3,4,3,2,6,12,5,1,2,2)
npchart <- qcc(x, type = "np", sizes = 500)
# try removing the limits
x <- x[-npchart$violations[1]$beyond.limits]
npchart2 <- qcc(x, type = "np", sizes = 500)
```

I would reccomend that the UCL = `r npchart2$limits[2]` and LCL = `r npchart2$limits[1]`.  Center line would be `r npchart2$center`.   These limits are calculated after having removed the OOC point and then recalculating the statistics.  

## 7.36  

```{r}
x <- c(1,0,4,3,1,2,5,0,2,1,1,0,8,0,2,1,3,5,4,6,3,1,0,2,4)
cchart <- qcc(x, type = "c")
# changing and computing a U chart
xc <- matrix(x, ncol=5)
x.new <- apply(xc, 1, sum)
#uchart <- qcc(x.new, type = "u", sizes = 5)
```

No.  The process is not in control.  There is one point that is beyond the upper control limit.  We can recompute the limits after removing the point that is OOC and report these new limits as the recommened ones.  

```{r}
x <- x[-cchart$violations$beyond.limits]
cchart2 <- qcc(x, type = "c")
```

Looking at this chart, the two yellow points are warnings per the Shewhart Rules, i.e. a run of 7 points below or above the center line.  Investigation may be warranted.  

## 7.41  

```{r}
x <- c(1,1,3,7,8,10,5,13,0,19,24,6,9,11,15,8,3,6,7,4,9,20)
cchart <- qcc(x, type = "c")

```

The process has 3 ooc points one of which is the most recent sample.  The process is NOT in statistical control.  

## 7.43  


## 7.44  

```{r}
# put options back to default
qcc.options(old)
```

