---
title: "PHYS 2210 Exam 2"
author: "Cody Frisby"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = F, message = F,
                      dpi = 200, fig.align = 200)
```

## 1.  Suppose a 900 kg satellite is orbiting the earth one earth radius above the surface.  

### a. **Find the force of gravity between the earth and the satellite**.  

```{r}
m1 <- 900 # mass of satellite (kg)
m2 <- 5.972e24 # mass of earth (kg)
r <- 6.371e6 # radius of earth (m)
G <- 6.67408e-11 # gravitational constant
FG <- (G * m1 * m2)/(2*r)^2

FGrav <- function(m1, m2, r) {
  G <- 6.67408e-11 # universal constant
  RESULT <- (G * m1 * m2) / r^2
  return(RESULT)
}
```

This can be found using 

$$F_g = G\frac{M_{satellite}M_{earth}}{r^2} = `r round(G,2)`\frac{(`r round(m1,2)`)(`r round(m2,2)`)}{`r round(2*r^2,2)`} = `r round(FG,2)` ~N$$  

where $r$ is the distance from the center of the earth to the satellite and $G$ is the gravitational constant.  

### b. **Find the orbital period of the satellite**.  

```{r}
velocity <- function(m, r) {
  G <- 6.67408e-11 # universal constant
  RESULT <- sqrt((G * m) / r)
  return(RESULT)
}
period <- function(m, r) {
  G <- 6.67408e-11 # universal constant
  RESULT <- 2 * pi * sqrt(r^3 / (G * m))
  return(RESULT)
}

p <- period(m2, 2*r)
v <- velocity(m2, 2*r)
```

This can be found using 

$$T = 2 \pi \sqrt{\frac {r^3} {GM_{earth}}} = 2\pi \sqrt{\frac {`r round(2*r^3,2)`} {`r round(G,2)` (`r round(m2,2)`)}} =`r round(p,2)` ~seconds$$

### c. **Find the tangential velocity required to keep the satellite in orbit**.  

This can be found using 

$$ v = \sqrt{\frac{GM_{earth}}{r}} = \sqrt{\frac{`r round(G * m2,2)`}{`round(r 2 * r,2)`}} = `r round(v,2)` ~\frac{m}{s}$$

### d. **Find the angular velocity of the satellite**.  

```{r}
angular <- function(m, r) {
  G <- 6.67408e-11 # universal constant
  RESULT <- sqrt((G * m) / r^3)
  return(RESULT)
}
a <- angular(m2, 2*r)
```

Angular velocity can be found using $\omega^2r = \frac{GM_E}{r^2}$ and solving for $\omega$ (where $\omega$ is the angular velocity) we get  

$$\omega = \sqrt {\frac {GM_{earth}} {r^3} } = \sqrt { \frac {`r round(G * m2,2)`} {`r round((2*r)^3,2)`} } = `r round(a,6)` ~ \frac{rad}{s}$$  

**Side note:**  We could find the orbital period from here since $$T = \frac{2 \pi}{\omega} = \frac{2 \pi}{`r round(a, 6)`} = `r  round(1 / (a / (2*pi)),2)`$$  which is the same answer we got for part **b**.  


## 2) A painter is standing on 4 kg board which is 12 m long and suspended on each end by a rope. The painter is standing 3 m from the left side, closer to the first rope. The mass of the painter is 90 kg.  

```{r}
m1 <- 4
m2 <- 90
l <- 12
x <- 3
g <- 9.8
t1 <- g * (m1*l/2 + m2*(l-x)) / l
```


### a. **Find the tension in the first rope**.  

The system is in equilibrium so the sum of the forces is zero ($\sum F = 0$) AND the sum of the torques is zero ($\sum \tau = 0$).  

$$(m_{board}(CM) + m_{painter}(L - x))g + TL $$ where $CM$ is the center of mass of the board and $L - x$ is the distance from the edge of the board (L) that the painter is standing on and $g$ is the acceleration of gravity.  So 

$$T_1 = g\frac {m_1 \frac{L}{2} + m_2(L-x)} {L} = `r t1` ~N$$

### b. **Find the tension in the second rope**.  

```{r}
t2 <- g * (m1 + m2) - t1
```


$$T_2 = (m_1 + m_2)9.8 - Tension_1 = `r t2` ~N$$  

Checking the equilibrium of the system, i.e. $g(m_1 + m_2) = T_1 + T_2$, we can see that $$9.8(`r m1` + `r m2`) = `r t1` + `r t2` = `r t1+t2`~N$$  


### c. **The painter climbs down from the board, just before the second rope suddenly snaps. Find the instantaneous angular acceleration of the board**.  

```{r}
i <- m1*l^2 / 3
tau <- m1*g*(l/2)
```

Since $L = I\omega$, where L is the angular momentum, angular acceleration can be found by $\tau = I\alpha$, where $\alpha$ is the angular acceleration, so angular acceleration can be found by $\alpha = \frac{\tau}{I}$.  

So we need to find I and $\tau$.  Using $I = \frac{Ma^2}{3} = `r i`$ since the pivot point is now 12 meters from where the rope snaps and $\tau = mg \frac{L}{2} = `r tau`$.    
$$\alpha = \frac{`r tau`}{`r i`} = `r tau/i` ~\frac{rad}{s}$$


## 3) A certain thin (one-dimensional) rod has a length of L and a non-uniform density which is given by the following function:  

$$\rho(L) = \frac{L^2}{7}$$  

We can picture the density of the rod as a function of L.  A plot of this function is pictured here for reference for lengths from 0 to 10.  

```{r}
rho <- function(x) (x^2) / 7
curve(rho(x), 0, 10, col = "dodgerblue", xlab = "L", 
      ylab = expression(rho(L)))
grid(); abline(h = 0, lty = 3); abline(v = 0, lty = 3)
```

We can see that the density takes on what appears to be an exponential function, meaning that the density of the rod increases exponentially with its length.  

### a. **Find the rod’s center of mass as a function of L**.  

The mass of our object depends on it's length.  

$$ m = \int_0^L\frac{1}{7}x^2 dx = \frac{1}{21} L^3 ~ kg $$

Now, the center or mass, or $\bar x$, can be defined as 
$$ \bar x = \int_0^L \frac{\rho(x) dx}{m}x$$ where I have already found $m$ above, I focus on the top piece.

$$ \bar x = \frac{{\int_0^L \rho(x)xdx}}{m} = \frac{1}{m} \int_0^L \frac {x^3} {7}dx = \frac {1} {28m} L^4 $$ 

Plugging in for m we get 

$$\bar x = \frac {21L^4} {28L^3} = \frac{3}{4}L $$

### b. **Find the moment of inertia for the rod if it is rotated about its midpoint**$(\frac{L}{2})$.  

In general the moment of inertia ($I$) can be defined as

$$I = mr^2$$   

where $m$ is the objects mass and $r$ is the distance of the object from the axis of rotation.  In part *a* we found that $m = \frac{1}{21}L^3$.  For a solid object we have to take all the particles that make up it's mass into account (you didn't make this easy).  Using (since it's one-dimensional linear density)

$$ dI = x^2 dm $$ I will attempt to find the moment of inertia of our non-uniform rod.  Recall that $\rho(x) = \frac{x^2}{7}$ (our density function) and $dm = \rho(x) dx$.  

$$dI = \frac{x^4}{7} dx $$

$$I = \int_{-\frac{L}{2}}^{\frac{L}{2}} dI = \int_{-\frac{L}{2}}^{\frac{L}{2}} \frac{x^4}{7} dx = \frac {1} {35}[\frac{L^5}{2^5} - (-\frac {L^5} {2^5})] = \frac{1}{560}L^5 $$  
**Note:** This is about 45% smaller than if we were to use the formula for the moment of inertia for a rod of uniform density($\frac{1}{12}MR^2$), which would end up being approx. $\frac{1}{252}L^5$.    

## 4) A large metal ball with a uniform density is dropped into a lake. The ball has a radius of 12 cm and (beginning at rest) sinks 40 m to the bottom in 3.6 s.  

### a. **How much water pressure does the ball experience at the bottom of the lake**?  

```{r}
r <- 12 / 100 # meters, radius of ball
h <- 40 # depth of lake bottom
t <- 3.6 # seconds it takes to sing
g <- 9.8 # we are on earth
vb <- (4/3) * pi * r^3
ab <- pi * r^2
p <- 1000 * 9.8 * 40
a <- 2 * h / (t^2)
pball <- round((g * 1000) / (9.8 - a), 2)
```

Using 

$$ Pressure = \frac{Force}{Area}$$ 

we can then rewrite this as 

$$P = \frac {\rho g h \pi r^2} {\pi r^2} = \rho g h $$

and applying it to our problem we get (assuming the density of the lake water is $1000 \frac{kg}{m^3}$)

$$P = 1000 \frac{kg}{m^3} (9.8)\frac{m}{s^2} (40)m =  `r p` Pa$$

### b. **What is the ball made of? (Hint: Find its density)**. 

Density ($\rho$) is equal to the volume of an object divided by it's mass.  

Ignoring the drag force, the net force ($F_{net}$) of the system is 
$$F_{net} = W_{ball} - F_{b} = 9.8 \rho_{ball}V - 9.8 \rho_{water} V$$ where $F_b$ is the buoyant force and $V$ is the volume of the ball.  

In order to solve this we need to know the acceleration of the ball (we could also use this equality to solve for $a$ if we knew the density of the ball and water).  Using the kinematic equation, $x = \frac{1}{2} a t^2$, I can solve for $a$ since I know $x$ and $t$.  

$$a = \frac{2x}{t^2} = \frac{80}{3.6^2} = `r round(a, 2)`~ \frac{m}{s^2}$$  

Doing some more rearranging I get 

$$\rho_{ball} = \frac {9.8 (1000) \frac{kg}{m^3}} {9.8 - `r a`} = `r round(pball, 2)` ~ \frac{kg}{m^3}$$  

The ball is made of aluminum (Al), which has a density of $2.70 \frac{g}{cm^3}$. 

## 5) Astronauts use springs to work out in space because there isn’t enough gravity to provide significant resistance to keep their muscles from atrophying. Suppose an astronaut is attached to a horizontal spring and begins oscillating back-and-forth once every 2 seconds. Back on Earth, a 20 kg mass was hung vertically on this same spring, causing it to stretch 21 cm.  

### a. **What is the mass of the astronaut**?  

```{r}
# spring constant
m1 <- 20 # kg
g <- 9.8 # gravity on earth
x <- 0.21 # m, distance the spring is stretched
k <- (m1 * g) / (x)
period <- 2
m2 <- (k * period^2) / (4 * pi^2)
```

First, the spring constant is 
$$F = -kx$$  

$$k = \frac{mg}{x} = \frac{(20 kg) (9.8) \frac{m}{s^2}}{0.21 m} = `r round(k,2)` ~\frac{N}{m}$$  

and now mass can now be found with period, $T = 2 s$.  The period is defined as 
$$T = 2\pi \sqrt {\frac {m} {k}}$$ 

and so solving for $m$ we get

$$m = \frac {k T^2} {4 \pi^2} = \frac {`r round(k,2)` (4)} {4 \pi^2} =  `r m2`~ kg$$  


## 6) Suppose a certain musical instrument consists of a pipe which is open at one end and closed at the other. Music is produced as sound waves move through the pipe, creating standing wave patterns. The pipe has a length of 80 cm.  

### a. **Find the wavelength of the first, third, and fifth harmonics (show your work)**.  

```{r}
freq <- function(n = 1, v = 343, l = 0.80) {
  RESULT <- (n * v) / (4 * l)
  return(RESULT)
}
wavelength <- function(n = 1, l = 0.80) {
  RESULT <- (4 * l) / (n)
  return(RESULT)
}
w1 <- wavelength(n = 1)
w3 <- wavelength(n = 3)
w5 <- wavelength(n = 5)
```

Using $$f_n = \frac{nv}{4L}$$ where $f_n$ is the nth frequency and $v$ is the speed of sound (I'll use $343 \frac{m}{s}$) and $L$ is the length of the pipe, in this case $0.8~m$. Wavelength, ($\lambda$), can be found using $$\lambda_n = \frac{4L}{n}$$ 

$$\lambda_1 = \frac {4 (0.80)} {1} = `r round(w1,2)`~m$$
$$\lambda_3 = \frac {4 (0.80)} {3} = `r round(w3,2)`~m$$
$$\lambda_5 = \frac {4 (0.80)} {5} = `r round(w5,2)`~m$$


### b. **Find the frequency of the first, third, and fifth harmonics (show your work)**.  

```{r}
f1 <- freq(n = 1)
f3 <- freq(n = 3)
f5 <- freq(n = 5)
```


$$f_1 = \frac {1 (343) \frac{m}{s}} {4 (0.80) m} = `r round(f1,2)`~Hz$$

$$f_3 = \frac {3 (343) \frac{m}{s}} {4 (0.80) m} = `r round(f3,2)`~Hz$$

$$f_5 = \frac {5 (343) \frac{m}{s}} {4 (0.80) m} = `r round(f5,2)`~Hz$$


### c. **Suppose you were riding your bike at 18 m/s towards the instrument as the fifth harmonic was being played. What frequency would you hear**?  

We can use what's known as the Doppler effect to solve this one.  This is stated in the equation 
$$f_L = \frac {\nu + v_L} {\nu + v_s}f_s$$ where $\nu$ is the speed of sound, $f_L$ is the frequency as heard by the listener, $v_L$ is the velocity of the listener, $f_s$ is the frequency of the source, and $v_s$ is the velocity of the source.  

```{r}
doppler <- function(v = 343, vl, vs, n = 1, l = 0.8) {
  RESULT <- ((v + vl) / (v + vs)) * freq(n, v, l)
}
fl <- doppler(vl = 18, vs = 0, n = 5)
```

$$f_L = \frac {(343 + 18) \frac {m}{s}} {(343 + 0)\frac{m}{s}} `r round(f5,2)`~Hz = `r round(fl,2)` ~ Hz$$  
This is slightly higher than the original pitch which is what we would expect.  

