---
title: "Enviornmental Impacts"
author: "Cody Frisby"
date: "March 29, 2016"
output: 
  html_document: 
    keep_md: yes
---

### I apologize for how messy this code is.  It was a work in progress while learning about some R functions.  If you have any questions about it please let me know.  codyfrisby@gmail.com  

```{r}
df <- read.table("~/Documents/MATH3710/enviornment/EnvironmentalImpacts.txt", header = TRUE)
# log variables
df$loghydro <- log(df$Hydrocarbons)
df$lognitro <- log(df$Nitrogen)
df$logSO2 <- log(df$SO2)
# create a var with a formula of the entire model
f <- formula(AAMort ~ .)
fit.all <- lm(f, data = df[1:16]) #not including log vars
summary(fit.all) # look at model
car::vif(fit.all) # variance inflation factors, car package needed
```

For the last function here, you will need the *car* pacakage installed if you don't already.  

```{r, dpi=200, fig.align='center'}
n.list <- c("AAMort", "AnnPrecip", "MeanJanTemp", 
            "MeanJulyTemp",
           "PctNonWhite", "PctWhiteCollar", "PctU20000", 
           "Hydrocarbons", "Nitrogen", "SO2", "RelHumid")
plot(df[n.list])
```


Now we will create another scatter plot matrix.

```{r, dpi=200, fig.align='center'}
n2 <- c('AAMort', 'AnnPrecip', 'MeanJanTemp', 'MeanJulyTemp', 'PctNonWhite', 'PctWhiteCollar', 'PctU20000', 'loghydro', 'lognitro', 'logSO2', 'RelHumid')
plot(df[n2])
```


Correlation Coefs:

```{r}
cor(df[1:16]) #I'm not including the added log vars
```


Fitting the model that is in the SAS example:

```{r}
fit1 <- lm(AAMort ~ ., df[!names(df) %in% 
                            c("Hydrocarbons", "SO2",
                              "Nitrogen", "loghydro")])
# Here we use R's vectorizing abilities to subset the 
# R object df. ! is the negating symbol, ie. "not" in blank.
summary(fit1)
```

Summary of the model above.  Now we plot the diagnostics.

```{r, dpi=200, fig.align='center'}
par(mfrow=c(1,2))
plot(fit1, which = c(1,2))
```

Now, we want to take a look at the variance inflation factors from this model.  

```{r}
car::vif(fit1)
```

Now, let's fit the next model from the SAS code example.

```{r}
# fit the model WITHOUT the 4 terms listed
fit2 <- lm(AAMort ~ ., df[!names(df) %in% 
                            c("Hydrocarbons", "SO2",
                              "Nitrogen", "lognitro")])
summary(fit2) # summary of the model
car::vif(fit2) # variance inflation factors
```

  
Model Diagnostics:  

```{r, dpi=200, fig.align='center'}
par(mfrow=c(1,2))
plot(fit2, which = c(1,2))
```

Histogram of the residuals from fit2 model.

```{r, dpi=200, fig.align='center'}
# function to draw a hist with a normal curve
draw.hist <- function(x, main = NULL) {
  h <- hist(
    x, breaks = 20, col = "lightblue", xlab = "", 
    main = main) 
  xfit <- seq(min(x),max(x), 0.01)
  yfit <- dnorm(xfit,mean = mean(x),sd = sd(x))
  yfit <- yfit * diff(h$mids[1:2]) * length(x)
  lines(xfit, yfit, col = "black", lwd = 2)
}
# above is a function, below we call the function and give it two 
# arguments.  Here, x is the residuals of the fit2 model.
draw.hist(x = fit2$residuals, main="Histogram of the Residuals")
```



N best models:
```{r}
# library MASS needed for funciton stepAIC
step.mods <- MASS::stepAIC(fit1, direction = "backward")
```

Notice, this function/algorithm automatically prints each "step" to the console.  It stops based on AIC criteria.  We can see a summary by

```{r}
step.mods$anova
```

Also, the "best" model is now subsetted into a data frame when we run

```{r}
new.df <- step.mods$model
```


And to fit this new, reduced "best" model

```{r}
fit3 <- lm(AAMort ~ ., data = new.df)
summary(fit3)
# remove PopPerHouse, fit model, and compare to the one with it
fit4 <- lm(AAMort ~ ., data = new.df[!names(new.df) %in% "PopPerHouse"])
anova(fit4, fit3)
```

So, our algorithm didn't remove PopPerHouse, but there are no adverse effects if we exclude it.  So, we will go with model **fit4**.

```{r}
summary(fit4)
```

Our new, reduced model diagnostic plots.  

```{r, dpi=200, fig.align='center'}
par(mfrow=c(1,2))
plot(fit4, which = c(1,2))
```

Now try to use the leaps package

## All Subsets  

Here is some code to select your TOP 30 models....You have to supply the x values (your predictors) in the form of a subsetted data frame or a matrix. 

```{r}
# here we get the top 30 models based on Cp.
# we could use adj.r2 or r2 as well
# we need to tell the leaps function what our x (predictors) are
# and our response, y, is.
mods.cp <- leaps::leaps(x = df[,c(1:11,15,17,19)], y = df[,16], 
            method="Cp", names = names(df[,c(1:11,15,17,19)]), nbest=30)
# number of models the above line genterated
length(mods.cp$Cp)
ord <- order(mods.cp$Cp) # orders the Cp values smallest to largest
top30 <- ord[1:30] # creates an indexing vector of length 30
mods30.cp <- mods.cp$which[top30, ]
# here we only used Cp, we could do the same for adjusted r square.
regs <- leaps::regsubsets(AAMort ~ AnnPrecip+MeanJanTemp+MeanJulyTemp+
                            PctGT65+PopPerHouse+School+PctSound+
                            PopPerSqMile+PctNonWhite+PctWhiteCollar+
                            PctU20000+RelHumid+loghydro+lognitro
                          +logSO2, data=df, 
                          nbest = 30)
tmp <- summary(regs)$outmat # matrix of our models
regs.data <- as.matrix(cbind(summary(regs)$rsq, summary(regs)$adjr2, 
                   summary(regs)$cp, summary(regs)$bic))
colnames(regs.data) <- c("r2", "adjr2", "cp", "bic")
# now we have a matrix with the above values
bic.ord <- order(regs.data[,4])
tmp30 <- cbind(1:30, tmp[bic.ord[1:30], ])
tmp30
```

And if we just want to look at the best model based on BICs and using the leaps package  

```{r}
tmp[bic.ord[1],]
```

Cool thing about this is we didn't need to exclude one of the log vars due to VIF.  BICs and other criteria selection methods take care of this for us.  

```{r}
library(leaps)
df.sub1 <- df[!names(df) %in% 
                c("Hydrocarbons", "SO2", "Nitrogen", "lognitro")]
leaps <- regsubsets(AAMort ~ ., data = df.sub1, nbest = 6)
#summary(leaps) #this produces very large output
```

Plot the model subsets  

```{r, dpi=200, fig.align='center', fig.height=8}
plot(leaps, scale = "r2")
```


Plot statistic by subset size 

```{r, dpi=200, fig.align='center'}
# can't knit this line.  Exclude it
#car::subsets(leaps, statistic="rsq")
```


Another way:

```{r}
fit.0 <- lm(AAMort ~ 1, data = df) # fit a model with no predictors
step(fit.0, scope = list(upper=fit1), data = df, direction = "both")
```

This way concludes a model with 6 terms, just like above.

