---
title: "Exam 1 STAT 4600 Take 2"
author: "Cody Frisby"
date: "October 7, 2016"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, dpi=200, fig.align = 'center', 
                      cache = F)
```

## Part B  

First, I display the population we will do our simulation and sampling from.  It is the standard normal distribution, $N(0,1)$

```{r normal hist}
curve(dnorm, col="lightblue", from = -4, to = 4, xlab="", ylab = "")
cord.x <- c(-4,seq(-4, 4, 0.01),4)
cord.y <- c(0,dnorm(seq(-4, 4, 0.01),0,1),0)
polygon(cord.x,cord.y, col="lightblue")
```

## Procedure:  

```{r normal}
# I simulate data from normal with mean=0 and sd=1
# I then take n samples and comple a t.test confidence interval
# and then bootrap n*b times and compute 95% quantile.
# we then compare the two results' accuracy. 
n <- 10; b <- 1000
x <- rnorm(n*b)
######### t test ##########
a <- matrix(sample(x, n*b, replace=TRUE), b, n)
f <- function(x){t.test(x)$conf.int} # create t conf intervals
d <- t(apply(a, 1, f)) # b t.test confidence intervals
y <- sign(d[,1]) * sign(d[,2]) # test for mu in interval
t.accuracy <- sum(y<0) / length(y) # how many contain mu?
######### end t test ###############
# what if we only have one sample from the normal?
######### bootstrap one random sample #############
w <- matrix(nrow=b, ncol=2)
for(i in 1:b){ # I can't figure this part out without a loop, so...
  B <- matrix(sample(a[i,], n*b, replace = TRUE), b, n)
  #bootstrap b samples n times, and b times in the loop
  v <- apply(B, 1, mean) # here we get 1000 means
  w[i,] <- quantile(v, c(0.025, 0.975)) # here we get the quantile...
}                                       # for the b means
z <- sign(w[,1]) * sign(w[,2]) # test to see if mu is in interval
boot.accuracy <- sum(z<0) / length(z) # how many times is mu there? 
```

I take $`r b`$ samples of size 10.  After the samples have been taken we calculate a 95% t test confidence interval for all $`r b`$ samples.  Of these $`r b`$ confidence intervals, I then determine how many of them contain $\mu$.  This is our accuracy measure.  

With the re-sampling procedure, we start with one sample of size 10 from the standard normal distribution.  I then re sample from this sample $`r b * n`$ times with replacement producing $`r b`$ samples of size `r n`.  Of these samples I then compute `r b` means and compute a 95% quantile.  We then repeat `r b` times.  After this is done I determine which quantiles contain $\mu$ and calculate that as a percent of the total.  Here are the results:   

$$t~test~accuracy = `r t.accuracy`$$  $$resample~accuracy = 
`r boot.accuracy`$$

-----  

A few thoughts about the above procedure:  We could just as easily do this procedure using different values for the quantiles and similarly for the t test confidence interval.  We could also do this procedure with sampling from other distributions.   


------

I now perform the same simulation as above, only changing the distribution to an exponential with rate = 1.  

```{r exponential}
n <- 10; b <- 1000
x <- rexp(n*b) # random samples from exponential distribution
######### t test ##########
a <- matrix(sample(x, n*b, replace=TRUE), b, n)
f <- function(x){t.test(x)$conf.int} # create t conf intervals
d <- t(apply(a, 1, f)) # b t.test confidence intervals
y <- d[,1] < 1 & d[,2] > 1 # test for mu in interval
t.accuracy <- sum(y) / length(y) # how many contain mu?
######### end t test ###############
# what if we only have one sample from the normal?
######### bootstrap one random sample #############
x <- matrix(nrow=b, ncol=2)
for(i in 1:b){ # I can't figure this part out without a loop, so...
  B <- matrix(sample(a[i,], n*b, replace = TRUE), b, n)
  #bootstrap b samples n times, and b times in the loop
  v <- apply(B, 1, mean) # here we get 1000 means
  x[i,] <- quantile(v, c(0.025, 0.975)) # here we get the quantile...
}                                       # for the b means
z <- x[,1] < 1 & x[,2] > 1 # test to see if mu is in interval
boot.accuracy <- sum(z) / length(z) # how many times is mu there?
```

Comparing the accuracy of the 95% *t test* confidence interval with the percent the bootstrap procedure gets correct our results are  $$t.test~accuracy = `r t.accuracy`$$  $$bootstrap~accuracy = `r boot.accuracy`$$  and surprisingly, the t test wins, even when the data is nowhere near normally distributed.  

-----

## Appendix I  

### R Code:
```{r code, eval=FALSE, echo=TRUE}
curve(dnorm, col="lightblue", from = -4, to = 4)
cord.x <- c(-4,seq(-4, 4, 0.01),4)
cord.y <- c(0,dnorm(seq(-4, 4, 0.01),0,1),0)
polygon(cord.x,cord.y, col="lightblue")
# simulate data from normal with mean=0 and sd=1
# then take n samples and comple a t.test confidence interval
# and then bootrap n*b times and compute 95% quantile.
# we then compare the two results' accuracy. 
n <- 10; b <- 1000
x <- rnorm(n*b)
######### t test ##########
a <- matrix(sample(x, n*b), b, n)
f <- function(x){t.test(x)$conf.int} # create t conf intervals
d <- t(apply(a, 1, f)) # b t.test confidence intervals
y <- sign(d[,1]) * sign(d[,2]) # test for mu in interval
t.accuracy <- sum(y<0) / length(y) # how many contain mu?
######### end t test ###############
# what if we only have one sample from the normal?
######### bootstrap one random sample #############
w <- matrix(nrow=b, ncol=2)
for(i in 1:b){ # I can't figure this part out without a loop, so...
  B <- matrix(sample(a[i,], n*b, replace = TRUE), b, n)
  #bootstrap b samples n times, and b times in the loop
  v <- apply(B, 1, mean) # here we get 1000 means
  w[i,] <- quantile(v, c(0.025, 0.975)) # here we get the quantile...
}                                       # for the b means
z <- sign(w[,1]) * sign(w[,2]) # test to see if mu is in interval
boot.accuracy <- sum(z<0) / length(z) # how many times is mu there?
t.accuracy # print to console
boot.accuracy # print to console
```

------  

## Appendix II  

Here I display a random sample of 40 of the confidence intervals produced with the bootstrap method for reference.  

`r knitr::kable(w[sample(1:1000, 40), ])`
