---
title: "NY Rivers"
author: "Cody Frisby"
date: "March 3, 2016"
output: 
  html_document: 
    keep_md: yes
---

Here we repeat the examples from class using the New York Rivers data.  You can find this csv file in the directory "MATH3710/datafiles".  

```{r, dpi=200, fig.align='center'}
ny <- read.csv("~/Documents/MATH3710/datafiles/nyriver.csv", header = T)
# get rid of the index var
ny <- ny[,2:7]
ny_reduced <- ny[!(ny$River %in% c("Neversink", "Hackensack", "Fishkill", 
                             "Oatka", "Wappinger")),]
# matrix scatterplot, exploratory step
plot(ny)
plot(ny_reduced)
```

In R, when you fit a linear model using lm(y ~ x1+x2+.....+xk) and store it in a variable, you have access to dffits, hatvals, and all sorts of these diagnostic values.  

```{r}
# fit a linear model
fit <- lm(Nitrogen ~ ComIndl, data = ny)
# store the hatvals
h <- hatvalues(fit)
# hat matrix, define X first
X <- as.matrix(cbind(rep(1), ny$ComIndl))
H <- X %*% solve(t(X) %*% X) %*% t(X)
H <- diag(H)
# diag(H) = h
# get the dfits
d <- dffits(fit)
# you can also do this
influence.measures(fit)
```

You're probably familiar with the diagnostic plots when you call plot(fit, which = c(1,2)).  When you plot a model, R has 4 built in plots that it returns.  Now, we'd like the other two as well.

```{r, dpi=200, fig.align='center'}
par(mfrow=c(2,2))
plot(fit)
```

From the 4th plot and the output above, we can identify observation 4 and 5 as influential points.

The reduced data set, fit.reduced, diagnostic plots:

```{r, dpi=200, fig.align='center'}
fit.reduced <- lm(Nitrogen ~ ComIndl, data = ny_reduced)
par(mfrow=c(2,2))
plot(fit.reduced)
```

The scatter plots from the original data set and the reduced one:

```{r, dpi=200, fig.align='center', fig.height=3}
library(ggplot2)
g <- ggplot(data=ny, aes(x=ComIndl, y=Nitrogen))
g <- g + geom_point(size = 4, colour = "black")
g <- g + geom_smooth(method = "lm", colour = "black")
gg <- ggplot(data=ny_reduced, aes(x=ComIndl, y=Nitrogen))
gg <- gg + geom_point(size = 4, colour = "black")
gg <- gg + geom_smooth(method = "lm", colour = "black")
gridExtra::grid.arrange(g,gg, ncol=2)
```

And the diagnostic data from the reduced data set.

```{r}
influence.measures(fit.reduced)
```

Just came across a few more R functions from various libraries that will assist you in model selcection and leverage/influential values.  

```{r, message=F, warning=F}
library(leaps)
# model selection
all.models <- regsubsets(Nitrogen ~ ., data = ny[,2:6])
all.models
leaps(x = ny[2:5], y = ny[,6])
```


```{r}
data(swiss)
a<-regsubsets(as.matrix(swiss[,-1]),swiss[,1])
summary(a)
b<-regsubsets(Fertility~.,data=swiss,nbest=2)
summary(b)
coef(a, 1:3)
vcov(a, 3)
```
