---
title: "Chapter 9 Homework"
author: "Cody Frisby"
date: "11/9/2016"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, fig.align = 'center', dpi=200, 
                      warning=F, message = F)
library(qcc); library(SixSigma); library(qualityTools)
old <- qcc.options()
qcc.options(bg.margin = "white")
## Assignment: ##
# 1, 2, 9, 18, 19
```


## 9.1  

```{r, fig.height=6}
x <- c(1045,1055,1037,1064,1095,1008,1050,1087,1125,1146,1139,1169,1151,
       1128,1238,1125,1163,1188,1146,1167)
mu <- 1050
s <- 25
k <- s/2
cs <- cusum(x, plot = TRUE, center = 1050, std.dev = 25, add.stats = F)
x.mu <- x - mu; C <- cumsum(x.mu)
xplus <- x - (mu + k)
xminus <- (mu - k) - x
cplus <- cs$pos
cminus <- cs$neg
df <- data.frame(x, x.mu, C, xplus, xminus, cplus, cminus)
#knitr::kable(df)
qc <- qcc(x, type = "xbar.one", plot = F)
s <- qc$std.dev
# for loop from Evan: 
n <- length(x)-1
v <- numeric(n+1)
v[1] <- 0
k <- .5
mu0 <- 10
for(i in 2:(n+1)) {
  v[i] <- max(0,x[i]-(mu0+k)+v[i-1])
}
w <- numeric(n+1)
w[1] <- 0.05
for(i in 2:(n+1)) {
  w[i] <- max(0, (mu0 - k) - x[i] + w[i-1])
}
```

K is typically 0.5 times the shift that is desired to be detected.  Since the desired detection is $1.0\sigma$  $$\frac{1}{2}\sigma = `r 25/2`$$.  

There are 11 observations that are beyond the boundry of $5\sigma = 125$.  


### b) 
The estimate for $\sigma$ in part a doesn't seem reasonable, $\hat \sigma = `r s`$.  

## 9.2  

Using $$y_i = \frac{x_i - \bar x}{\hat \sigma}$$  

```{r, fig.height=6}
x <- c(1045,1055,1037,1064,1095,1008,1050,1087,1125,1146,1139,1169,1151,
       1128,1238,1125,1163,1188,1146,1167)
x <- (x - mean(x))/sd(x)
qc <- qcc(x, type = "xbar.one", plot = F)
mu <- qc$center; s <- qc$std.dev
cs <- cusum(x, plot = T, add.stats = F)
x.mu <- x - mu; C <- cumsum(x.mu)
xplus <- x - (mu + k)
xminus <- (mu - k) - x
cplus <- cs$pos
cminus <- cs$neg
df <- data.frame(x, x.mu, C, xplus, xminus, cplus, cminus)
```

Standardizing has no effect on the ability to detect OOC points. 

## 9.9  

```{r}
x <- c(3169,3173,3162,3154,3139,3145,3160,3172,3175,3205,3203,3209,
       3208,3211,3214,3215,3209,3203,3185,3187,3192,3199,3197,3193,
       3190,3183,3197,3188,3183,3175,3174,3171,3180,3179,3175,3174)
qc <- qcc(x, type = "xbar.one", plot = F)
d2 <- 1.128
R <- abs(diff(x))
s <- mean(R)/d2
```

To estimate the process standard deviation, where $n = 1$, we first calculate the moving range of the oberservations $x$.  $$Moving~Range = `r R`$$  $$\hat \sigma = \frac{\bar R}{d_2} = \frac{`r mean(R)`}{`r d2`} = `r s`$$  


See plot below (red points are above or below h):  

```{r, comment= ""}

# for loop from Evan: 
n <- length(x)
v <- numeric(n)
v[1] <- 0
k <- 0.25
h <- 8.01
mu0 <- 3200
for(i in 2:(n)) {
  v[i] <- max(0, x[i] - (mu0+k) + v[i-1])
}
w <- numeric(n)
w[1] <- 30.75
for(i in 2:n) {
  w[i] <- max(0, (mu0 - k) - x[i] + w[i-1])
}
plot(c(1, length(x)) ,range(max(c(5, v)), min(-5, w*-1)), 
     type = "n", xlab = "", ylab = "")
points(v, type = "b", pch = 16, col = ifelse(v > h, "red", "black"))
points(w*-1, type = "b", pch = 16, col = ifelse(w*-1 < -h, "red", "black"))
abline(h = 0, lty = 1, lwd = 2)
abline(h = c(-1,1)*h, lty = 2)
```


## 9.18  

```{r}
u0 <- 10; s <- 1; i <- 1
# part a
lam <- 0.1; L <- 3
ucl <- u0 + L * s * sqrt((lam/(2-lam)) * (1 - (1-lam)^(2*i)))
lcl <- u0 - L * s * sqrt((lam/(2-lam)) * (1 - (1-lam)^(2*i)))
```

With $\lambda = `r lam`$ when $i = 1$

$$UCL = `r ucl`$$
$$LCL = `r lcl`$$

```{r}
# part b
lam <- 0.2; L <- 3
ucl <- u0 + L * s * sqrt((lam/(2-lam)) * (1 - (1-lam)^(2*i)))
lcl <- u0 - L * s * sqrt((lam/(2-lam)) * (1 - (1-lam)^(2*i)))
```

With $\lambda = `r lam`$ when $i = 1$

$$UCL = `r ucl`$$
$$LCL = `r lcl`$$

```{r}
# part c
lam <- 0.4; L <- 3
ucl <- u0 + L * s * sqrt((lam/(2-lam)) * (1 - (1-lam)^(2*i)))
lcl <- u0 - L * s * sqrt((lam/(2-lam)) * (1 - (1-lam)^(2*i)))
```

With $\lambda = `r lam`$ when $i = 1$

$$UCL = `r ucl`$$
$$LCL = `r lcl`$$

It appears that the affect of $\lambda$ is a positive one, meaning as $\lambda$ increases so do our LCL and UCL.  

As $i$ increases our control limits become steady.  For example, when $i = 20$ and $\lambda = `r lam`$ we get

```{r}
i <- 20
ucl <- u0 + L * s * sqrt((lam/(2-lam)) * (1 - (1-lam)^(2*i)))
lcl <- u0 - L * s * sqrt((lam/(2-lam)) * (1 - (1-lam)^(2*i)))
```

$$UCL = `r ucl`$$
$$LCL = `r lcl`$$

```{r}
i <- 20; lam = 0.2
ucl <- u0 + L * s * sqrt((lam/(2-lam)) * (1 - (1-lam)^(2*i)))
lcl <- u0 - L * s * sqrt((lam/(2-lam)) * (1 - (1-lam)^(2*i)))
```

Or with $\lambda = `r lam`$ we get 
$$UCL = `r ucl`$$
$$LCL = `r lcl`$$  


## 9.19  

```{r, fig.height=6}
# data from 9.4 with lam = 0.2, L = 3
x <- c(8,8.01,8.02,8.01,8,8.01,8.06,8.07,8.01,8.04,8.02,8.01,8.05,
       8.04,8.03,8.05,8.06,8.04,8.05,8.06,8.04,8.02,8.03,8.05)
qc <- qcc(x, type = "xbar.one", plot = F)
cc <- ewma(x, center = 8.02, std.dev = 0.05)
```

With $center = 8.02$ and $\hat \sigma = 0.05$, the process is in control.  After approximately point 11 or 12 the process limits are stable and $$UCL = `r cc$limits[24,2]`$$ 
$$LCL = `r cc$limits[24,1]`$$  

