---
title: "Homework 01"
author: "Cody Frisby"
date: "1/10/2017"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, fig.align = 'center', dpi=200,warning=F,message=F)
library(MVA) # library for the textbook.
```

## 1.1  

Here I display the correlation matrix is a little different way.  

```{r}
df <- read.csv("~/Documents/STAT4400/data/hypo.csv")
df <- df[-1]
n <- sapply(df, class) == "integer" | sapply(df, class) == "numeric" 
cov.mat <- cov(df[n], use = "pairwise.complete.obs")
cor.mat <- cor(df[n], use = "pairwise.complete.obs")
# or even better :)
library(ggplot2); library(GGally)
ggpairs(df[, c("age", "IQ", "weight")])
```

And here I display the covariance matrix. `r knitr::kable(cov.mat)`

## 1.2  

```{r}
hypo.update <- df
hypo.update$age <- ifelse(is.na(df$age), 
                          mean(df$age, na.rm = TRUE), df$age)
hypo.update$IQ <- ifelse(is.na(df$IQ), 
                          mean(df$IQ, na.rm = TRUE), df$IQ)
cov.mat <- cov(hypo.update[n])
cor.mat <- cor(hypo.update[n])
```

Covariance matrix: `r knitr::kable(cov.mat)`

Correlation matrix:  `r knitr::kable(cor.mat)`  


## 1.3  

Displaying the individual normality plots here:

```{r, fig.height=6, fig.width=7}
df <- read.csv("~/Documents/STAT4400/data/pottery.csv")
x <- df[-1]
par(mfrow = c(3,3))
for(i in 1:9){
qqnorm(x[,i], main = names(x)[i]); qqline(x[,i])  
}
```

And the $\chi^2$ plot here:

```{r}
row.names(x) <- df$X # to be able to display which obs are outliers on the plot
# following the example from page 19 in the textbook.
cm <- colMeans(x)
S <- cov(x)
d <- apply(x, 1, function(x) t(x-cm) %*% solve(S) %*% (x-cm))
plot(qc <- qchisq((1:nrow(x) - 1/2) / nrow(x), 9), sd <- sort(d), 
     xlab = expression(paste(chi[9]^2, " Quantile")), 
     ylab = "Ordered Distances", ylim = c(0, max(d)))
abline(0, 1, lty = 3, col = "red", lwd = 2)
outliers <- which(rank(abs(qc - sd), ties.method = "random") > nrow(x) - 3)
text(qc[outliers], sd[outliers] - 1.5, names(outliers)) # label the "outliers"
```

Where $$d_i^2 = (\mathbf{x_i} - \bar x)^T \mathbf{S}^{-1} ((\mathbf{x_i} - \bar x))$$

## 1.4  

```{r}
x <- matrix(c(3.8778,2.8110,3.1480,3.5062,
              2.8110,2.1210,2.2669,2.5690,
              3.1480,2.2669,2.6550,2.8341,
              3.5062,2.5690,2.8341,3.2352), byrow = T, ncol = 4)
# the way I did it before figuring out the book's syntax
d <- sqrt(diag(x))
y <- outer(d, d, "*")
cor.mat <- x/y
# the way the book does it 
d <- diag(x)
m <- diag(1/sqrt(d)) %*% x %*% diag(1/sqrt(d))
# running identical(round(m, i), round(cor.mat, i))
# for i in 1:16 all are TRUE except to 16 digits, so both methods 
# produce the same result, basically.
```
The correlation matrix is displayed here:
`r knitr::kable(cor.mat)`


## 1.5  

Using the R function 

```{r}
x <- matrix(c(3,4,4,6,1,5,1,1,7,3,6,2,0,2,6,1,1,1,0,3,4,7,3,6,2,2,2,
              5,1,0,0,4,1,1,1,0,6,4,3,5,7,6,5,1,4,2,1,4,3,1), ncol = 5)
d <- dist(scale(x, center = FALSE))
library(reshape2)
df <- melt(as.matrix(d), varnames = c("row", "col"))
```

The *Euclidean distances* for the data are `r knitr::kable(df[df$row > df$col, ])`  

```{r}
d <- dist(scale(x, center = FALSE), method = "manhattan")
df <- melt(as.matrix(d), varnames = c("row", "col"))
```

and the city block, or `manhattan`, distances for the data are 

`r knitr::kable(df[df$row > df$col, ])` 

## 1.6  

```{r}
df <- data.frame(recieptID = c(101,521,746,857), books = c(4,5,4,3), 
                 sales = c(142,252,148,158), 
                 student = c("yes", "yes", "no", "yes"))
```

### a)  

There are 4 observations, or units, in the data set, and there are 4 variables -  *reciept ID, number of books, dollars of sales*, and whether the customer was a *student* or not.  There are categorical and continuous types of variables in the data set.  Of the categorical set, *students*, there are two levels: **yes** and **no**.  

### b)  

`r knitr::kable(df)`

### c)  
Here I display rows 2 and 4 of the data frame above.

$$X_2 = `r df[2, ]`$$
$$X_4 = `r df[4, ]`$$

## 1.7  

```{r}
x <- c(1,1,1,1)
xbar <- c(.766, .508, .438, .161)
S <- matrix(c(.856,.635,.173,.096,.635,.568,.127,.067,.173,.128,
              .171,.039,.096,.067,.039,.043), ncol = 4)
v <- sum(diag(S)) + 2*sum(S[lower.tri(S)])
v.total <- t(x) %*% S %*% x
```

### a)  

For reference:  $x_1 = petroleum, ~ x_2 = natural~gas, ~ x_3 = hydroelectric~power, ~ x_4 = nuclear~electric~power.$  

The mean of a state's total energy consumption is simply the sum of the means of the 4 energy sources, $\sum \mathbf{\bar x_i} = `r sum(xbar)`$. The variance of a state would be the sum of the variances plus 2 times the covariance terms. 

$$\sigma_1^2 + \sigma_2^2 + \sigma_3^2 + \sigma_4^2 + 2cov_{12} + 2cov_{13} +...+2cov_{34} = `r v`$$ or using matrix notation 
$$\mathbf{x}^T \mathbf{S} \mathbf{x}$$ where x is a unit vector of length 4.  

### b)  

We are asked to find the mean of $x_1 - x_2$ as well as the $cov_{12}$.  

$$x_1 - x_2 = `r xbar[1]` - `r xbar[2]` = `r xbar[1] - xbar[2]`$$  

We can find the covariance of these two variables by taking the 2nd row 1st column term from the supplied matrix, S, $cov_{12} = `r S[2,1]`$.  

### c)  $Y_1 = X_1 + X_2$ and $Y_2 = X_1 - X_2$  

```{r}
m.y1 <- xbar[1] + xbar[2]
m.y2 <- xbar[1] - xbar[2]
s.y1 <- S[1,1] + S[2,2] + 2 * S[2,1]
s.y2 <- S[1,1] + S[2,2] - 2 * S[2,1]
cv <- S[1,1] - S[2,2]
m <- matrix(c(s.y1,cv ,cv ,s.y2), ncol = 2)
```


The variance is $`r S[1,1]` + `r S[2,2]` + 2\times`r S[2,1]` = `r s.y1` =  \sigma_{Y_1}^2$. And $\sigma_{Y_2}^2 = `r S[1,1]` + `r S[2,2]` - 2\times`r S[2,1]` = `r s.y2`$.  

And the variance-covariance matrix would be `r knitr::kable(m)`

------------------

## R Code:

```{r, eval=F, echo=T}
### 1.1
df <- read.csv("~/Documents/STAT4400/dat/hypo.csv")
df <- df[-1]
n <- sapply(df, class) == "integer" | sapply(df, class) == "numeric" 
cov.mat <- cov(df[n], use = "pairwise.complete.obs")
cor.mat <- cor(df[n], use = "pairwise.complete.obs")
# or even better :)
library(ggplot2); library(GGally)
ggpairs(df[, c("age", "IQ", "weight")])
### 1.2 
hypo.update <- df
hypo.update$age <- ifelse(is.na(df$age), 
                          mean(df$age, na.rm = TRUE), df$age)
hypo.update$IQ <- ifelse(is.na(df$IQ), 
                          mean(df$IQ, na.rm = TRUE), df$IQ)
cov.mat <- cov(hypo.update[n])
cor.mat <- cor(hypo.update[n])
### 1.3
df <- read.csv("~/Documents/STAT4400/dat/pottery.csv")
x <- df[-1]
par(mfrow = c(3,3))
for(i in 1:9){
qqnorm(x[,i], main = names(x)[i]); qqline(x[,i])  
}
row.names(x) <- df$X # to be able to display which obs are outliers on the plot
# following the example from page 19 in the textbook.
cm <- colMeans(x)
S <- cov(x)
d <- apply(x, 1, function(x) t(x-cm) %*% solve(S) %*% (x-cm))
plot(qc <- qchisq((1:nrow(x) - 1/2) / nrow(x), 9), sd <- sort(d), 
     xlab = expression(paste(chi[9]^2, " Quantile")), 
     ylab = "Ordered Distances", ylim = c(0, max(d)))
abline(0, 1, lty = 3, col = "red", lwd = 2)
outliers <- which(rank(abs(qc - sd), ties.method = "random") > nrow(x) - 3)
text(qc[outliers], sd[outliers] - 1.5, names(outliers)) # label the "outliers"
### 1.4 
x <- matrix(c(3.8778,2.8110,3.1480,3.5062,
              2.8110,2.1210,2.2669,2.5690,
              3.1480,2.2669,2.6550,2.8341,
              3.5062,2.5690,2.8341,3.2352), byrow = T, ncol = 4)
d <- sqrt(diag(x))
y <- outer(d, d, "*")
cor.mat <- x/y
### 1.5 
x <- matrix(c(3,4,4,6,1,5,1,1,7,3,6,2,0,2,6,1,1,1,0,3,4,7,3,6,2,2,2,
              5,1,0,0,4,1,1,1,0,6,4,3,5,7,6,5,1,4,2,1,4,3,1), ncol = 5)
d <- dist(scale(x, center = FALSE))
library(reshape2)
df <- melt(as.matrix(d), varnames = c("row", "col"))
d <- dist(scale(x, center = FALSE), method = "manhattan")
df <- melt(as.matrix(d), varnames = c("row", "col"))
### 1.6  
df <- data.frame(recieptID = c(101,521,746,857), books = c(4,5,4,3), 
                 sales = c(142,252,148,158), 
                 student = c("yes", "yes", "no", "yes"))
### 1.7 
xbar <- c(.766, .508, .438, .161)
S <- matrix(c(.856,.635,.173,.096,.635,.568,.127,.067,.173,.128,.171,.039,
              .096,.067,.039,.043), ncol = 4)
v <- sum(diag(S)) + 2*sum(S[lower.tri(S)])
m.y1 <- xbar[1] + xbar[2]
m.y2 <- xbar[1] - xbar[2]
s.y1 <- S[1,1] + S[2,2] + 2 * S[2,1]
s.y2 <- S[1,1] + S[2,2] - 2 * S[2,1]
cv <- S[1,1] - S[2,2]
m <- matrix(c(s.y1,cv ,cv ,s.y2), ncol = 2)
```

