---
title: "Quiz 8"
author: "Cody Frisby"
date: "11/4/2016"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, fig.align = 'center', dpi=200, 
                      warning=F, message = F)
library(qcc); library(SixSigma)
old <- qcc.options()
qcc.options(bg.margin = "white")
```


## 8.26  

```{r}
y <- read.csv("~/Documents/STAT4600/hw/hw06/8-26.csv")
y <- y$Ex8.32Reading
df <- data.frame(y, part = gl(10, 3), 
                 operator = factor(c(rep(1, 30), rep(2, 30))))
fit <- aov(y ~ part*operator, df)
#qc <- ss.rr(y, part, operator, data = df, main = "", lsl = 40, usl=60)
```

First I display the ANOVA table with the full model 
$$y_{ijk} = \mu + P_i + O_j + (PO)_{ij} + \varepsilon_{ijk}\left\{
        \begin{array}{ll}
            i = 1, 2, ..., 10 \\
            j = 1, 2 \\
            k = 1, 2, 3
        \end{array}
    \right.
    $$
Where $P_i$ is the ith part and $O_j$ is the jth operator.
`r knitr::kable(anova(fit))`  

It can be seen that the interaction term (as well as the operator term) is not significant.  I suggest we fit a reduced model that drops this term.  

$$y_{ijk} = \mu + P_i + O_j + \varepsilon_{ijk}\left\{
        \begin{array}{ll}
            i = 1, 2, ..., 10 \\
            j = 1, 2 \\
            k = 1, 2, 3
        \end{array}
    \right.
    $$

```{r}
fit2 <- aov(y ~ part + operator, df)
```

and the ANOVA table from the reduced model is shown here.  

`r knitr::kable(anova(fit2))`

```{r}
library(dplyr)
y.i <- df %>%
  group_by(part) %>%
  summarise_each(funs(sum), y)
y.j <- df %>%
  group_by(operator) %>%
  summarise_each(funs(sum), y)
y.. <- sum(df$y)
a <- dim(y.i)[1]; b <- dim(y.j)[1]; n <- 3 # number of replicates
N <- a*b*n
ssa <- (1/(b*n))*sum(y.i$y^2) - y..^2/N 
ssb <- (1/(a*n))*sum(y.j$y^2) - y..^2/N
sst <- sum(y^2) - y..^2/N
sse <- sst - ssa - ssb
msa <- ssa/(a-1); msb <- ssb/(b-1)
mse <- sse/(N-a-b+1)
s.p <- (msa - mse)/(b*n)
s.o <- ifelse((msb - mse)/(a*n)< 0, 0, (msb - msab)/(a*n))
#qc <- ss.rr(y, part, operator, data = df, main = "")
```
 
With the reduced model, we can estimate gauge repeatibility with $\hat \sigma^2$ wich is simplly the mean square error of our model, **`r mse`**.  

We can estimate gauge reproducibility with $$\sigma_{Reproducibility}^2 = \sigma_O^2 = 0$$  where $$\sigma_O^2 = \frac{MS_{operator} - MS_{Error}}{an}$$ and a = `r a` and n = `r n`, and if it is less than zero we assume it to be zero.  

```{r}
qcc.options(old)
```
