---
title: "Quiz from Chapter 9"
author: "Cody Frisby"
date: "11/11/2016"
output: pdf_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, fig.align = 'center', dpi=200, 
                      warning=F, message = F)
library(qcc); library(SixSigma); library(qualityTools)
old <- qcc.options()
qcc.options(bg.margin = "white")
## Quiz ##  
# number 1
```

A quick glance at the scatter plot of the observations would indicate a trend upward in the data.  This upward trend is picked up in the cummulative sum chart after around the 9th observation (see plot below).  

```{r, fig.height=6}
x <- c(1045,1055,1037,1064,1095,1008,1050,1087,1125,1146,1139,1169,1151,
       1128,1238,1125,1163,1188,1146,1167)
cs <- cusum(x, plot = TRUE, center = 1050, std.dev = 25, add.stats = F,
            title = "", xlab = "", ylab = "")
qc <- qcc(x, type = "xbar.one", plot = F)
s <- qc$std.dev
```


K is typically 0.5 times the shift that is desired to be detected.  Since the desired detection is $1.0\sigma$ 
$$\frac{1}{2}\sigma = `r 25/2`$$ $$H = 5\sigma = 125$$


### b) 
The estimate for $\sigma$ in part a doesn't seem reasonable, $\hat \sigma = `r s`$.  

------

The observations and plotted data points are 

```{r}
dt <- cbind(x, cs$pos, cs$neg)
colnames(dt) <- c("xi","Cplus", "Cnegative")
knitr::kable(dt)
```


```{r}
qcc.options(old)
```

