---
title: "Class Notes from 4710"
output:
  html_document:
    keep_md: yes
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, dpi=200, fig.align = 'center')
```

## Birthday problem from class 

```{r}
birthday <- function(n){(choose(365,n) * factorial(n)/(365^n))}
birthday(30)
```


```{r}
# coffee example from page 54.
library(gtools)
S <- permutations(3, 3, letters[24:26])
A <- S[S %in% "xyz" | S %in% "xzy" | S %in% "zxy"]
B <- S[S %in% "xyz" | S %in% "xzy"]
C <- S[S %in% "yxz" | S %in% "zxy"]
D <- S[S %in% "yzx" | S %in% "zyx"]
p.A <- length(A)/length(S); p.B <- length(B)/length(S)
p.C <- length(C)/length(S); p.D <- length(D)/length(S)
```

## Chapter.Section 3.2,  example 3.1

```{r Example 3.1}
# y can take on values of 0, 1, or 2 ONLY
p.y <- function(y){
  choose(3, y) * choose(3, 2 - y) * (1/choose(6,2))
}
p.y(0:2)

# Number 3.6, page 90
A <- combn(1:5, 2) # lists all the possible draws from the urn
A
# part A
for(i in 2:5){
  l <- length(which(A[2,] == i))
  print(l/10) #probs for 2, 3, 4, 5 respectively
}
# part B
s <- apply(A, 2, sum) # all possible sums
s
for(i in 3:9){
  l <- length(which(s == i))
  names(l) <- paste0("P(Y=",i,")")
  print(l/10)
}
```


```{r Exercise 3.7}
# 3.7, 3 balls & 3 bowls, probability of one empty bowl
library(gtools)
s <- permutations(3, 3,v=0:2, repeats.allowed = F) 
s # possibilities where there is one empty bowl
```


```{r 3.9}
# 3.9, accounting error and auditor
library(gtools)
s <- permutations(2, 3, letters[c(5,14)], repeats.allowed = TRUE)
s # possible sample space where "e"=error and "n"=no error
# 3.10, we can write a function for this one
# we start right after a rental happened.  So the probability
# of renting on the first day after a rental occured is 0.2
p.y <- function(y){
  0.2*0.8^y
}
p.y(0:10)
plot(0:10, p.y(0:10), type = "h", col="blue", lwd=3)
```

## 3.3 The Expected Value of a Random Variable or a Function of a Random Variable  

### Definition 3.4:

Let Y be a **discrete** random variable... 

$$E(Y) = \sum_{y}yp(y)$$

### Theorem 3.2:  

$$E[g(Y)] = \sum_{all y}g(y)p(y)$$  

### Definition 3.5: 

$$V(Y) = E[(Y - \mu^2)] = E[Y^2] - 2 \mu E[Y] + \mu^2 = E[Y^2] - \mu^2$$  

```{r}
E <- function(y,p.y){sum(y*p.y)} #E[Y]
V <- function(y,p.y){sum(y^2*p.y) - sum(y*p.y)^2} #E[V]
# Example 3.2, page 94
y <- 0:3
p.y <- c(1/8, 1/4, 3/8, 1/4)
mu <- sum(y*p.y)
mu
variance <- sum((y-mu)^2*p.y)
variance
sig <- sqrt(variance)
sig
# a plot of the data
plot(y, p.y, type="h", ylim=c(0,0.5), xlim = c(-0.5, 4), lwd=10, col="blue")
```

```{r Example 3.3}
# using Theorem 3.6 to find the variance of Y from example 3.2
sum(y^2*p.y) - mu^2 
identical(sum(y^2*p.y) - mu^2, variance) #same result as variance
```

### Example 3.4 

```{r Example 3.4}
Ca <- function(t){13*t + .3*t^2}
Cb <- function(t){11.6*t + .432*t^2}
Ca(c(10, 20))
Cb(c(10, 20))
```

```{r 3.12}
# 3.12
y <- 1:4 #values for Y
p.y <- seq(0.4, 0.1, by=-0.1) #p(Y)
sum(y * p.y) #E[Y]
sum(y^2*p.y) - 1 #E[Y^2 - 1]
sum((1/y)*p.y) #E[1/Y]
sum(y^2 * p.y) - sum(y * p.y)^2 #E[V]
```


```{r 3.13}
# 3.13
y <- c(1,2,-1)
p.y <- c(0.25, 0.25, 0.5)
sum(y * p.y) #E[Y]
sum(y^2*p.y) - sum(y*p.y)^2 #E[V]
```

```{r 3.14}
# 3.14
y <- 3:13
p.y <- c(.03,.05,.07,.1,.14,.2,.18,.12,.07,.03,.01)
sum(y * p.y) #E[Y]
sqrt(sum(y^2*p.y) - sum(y*p.y)^2) #E[sqrt(Y)]
sum(p.y) - (p.y[1] + p.y[11]) #p(2sd < Y < 2sd)
```

```{r 3.17}
# 3.17
y <- 0:2
p.y <- c(6/27, 18/27, 3/27)
E(y, p.y); sqrt(V(y,p.y)) #E[Y] & sqrt(E[V])
E(y,p.y) + 1.133115
```

```{r 3.21}
# 3.21
y <- 21:26
p.y <- c(.05,.2,.3,.25,.15,.05)
E(y,p.y)
V(y,p.y)
8*pi*E(y^2, p.y)
```

```{r 3.23}
# 3.23
y <- c(15,5,-4)
p.y <- c(8/52, 8/52,36/52)
E(y,p.y)
```

```{r 3.25}
# 3.25
permutations(3, 2, repeats.allowed = T) #sample space
```

## 3.4  Binomial Distribution
```{r 3.38}
# 3.38
y <- 3:4
sum(dbinom(y, 4, 1/3))
# or
sum(choose(4,y) * (1/3)^y * (2/3)^(4-y)) # same result
```

```{r 3.39}
# 3.39
y <- 0:2
sum(dbinom(y, 4, .2))
# or
sum(choose(4,y) * (.2)^y * (.8)^(4-y)) # same result
```

```{r 3.41}
# 3.41
y <- 10:15
sum(dbinom(y, 15, 1/5))
# or
sum(choose(15,y) * (1/5)^y * (4/5)^(15-y)) # same result
```


```{r 3.42}
# 3.42 part A
y <- 10:15
sum(dbinom(y, 15, 1/4))
# or
sum(choose(15,y) * (1/4)^y * (3/4)^(15-y)) # same result
# part B
sum(dbinom(y, 15, 1/3))
# or
sum(choose(15,y) * (1/3)^y * (2/3)^(15-y)) # same result
```

```{r 3.46, fig.height=3, fig.width=7}
# 3.46
x <- 0:5
par(mfrow=c(1,3))
plot(x,dbinom(0:5, 5, .1), lwd=2, col="blue", type = "h", main = "p=0.1")
plot(x,dbinom(0:5, 5, .5), lwd=2, col="blue", type = "h", main = "p=0.5")
plot(x,dbinom(0:5, 5, .9), lwd=2, col="blue", type = "h", main = "p=0.9")
```

```{r 3.47}
# 3.47
plot(0:20, dbinom(0:20, 20, 0.5), col="blue", lwd=3, type="h", 
     main = "p=0.5, n=20")
```


```{r 3.7}
# writing a hypergeometric function
h <- function(y,r,n,N){
  (choose(r,y) * choose(N-r, n-y)) / choose(N,n) 
}
```

## Chapter 4  


```{r}
## Exercise 4.18  
# f is the pdf, g is the cdf
f <- function(x){ # this is the probability density function
return(ifelse(x <= -1, 0, 
        ifelse((x > -1) & (x <=0), 0.2, 
              ifelse((x > 0) & (x <= 1), 0.2 + 1.2 * x, 0))))
  }
g <- function(x){ # this is the CDF
return(ifelse(x <= -1, 0, 
       ifelse((x > -1) & (x <=0), 0.2*(1 + x), 
              ifelse((x > 0) & (x <= 1), 0.2 * (1 + x + 3 * x^2), 1))))
} 
curve(f(x), -1, 1, lty=3, col="red", ylab="") # plot the pdf
curve(g(x), -1, 1, col="blue", add=TRUE) # plot the cdf
legend("topleft", legend = c("f(x)", "F(x)"), lty=c(3,1), # add a legend
       col=c("red", "blue"), bty="n")
```


## 5.2  

```{r, fig.height=5}
# Figure 5.1
library(plotly)
df <- data.frame(x = rep(1:6, 6), y = c(rep(1,6), rep(2,6), rep(3,6), 
                                        rep(4,6), rep(5,6), rep(6,6)))
df$z <- rep(1/36, 36)
plot_ly(df, x = ~x, y = ~y, z = ~z)
# the dots are at a height of 1/36

```


```{r}
# Example 5.1  
library(gtools)
a <- permutations(3, 2, 0:2, repeats.allowed = T)
b <- permutations(3, 2, 1:3, repeats.allowed = T)
```

```{r}
# normal distribution function: 
f <- function(x, mu=0, s=1){
  (1/(s*sqrt(2*pi))) * exp(-(1/(2*s^2))*(x - mu)^2)
}
integrate(f, -1, 1)$value # probability that x is between -1 and 1
1 - (pnorm(-1) + 1 - pnorm(1)) # same as above
```

## 4.6 Gamma Distribution  

```{r}
# exercise 4.84  
curve(dgamma(x, shape=4, scale=1/1), 0, 200, n=1000, ylab="f(x)",
      main = "Gamma Density Distributions")
curve(dgamma(x, shape=40, scale=1/1), add=TRUE, col="red", n=1000)
curve(dgamma(x, shape=80, scale=1/1),add=TRUE, col="blue", n=1000)
```

### Chi-square Distribution

```{r}
curve(dchisq(x, 1), ylab = "f(x)", main = "Chi-Square Density Distributions",
      xlim=c(0, 25))
curve(dchisq(x, 5), add = TRUE, col = "blue")
curve(dchisq(x, 15), add = TRUE, col = "red")
```

### Exponential Distribution  

```{r}
curve(dexp(x,1), 0, 10, col = "blue", ylab = "f(x)",
      main = "Exponential Density Functions")
curve(dexp(x,1/2), add=TRUE, col="red")
curve(dexp(x,1/15), add=TRUE, col="green")
```


## 4.7 Beta Distribution  

```{r}
curve(dbeta(x,2,2), 0, 1, col = "blue", ylim=c(0, 2.5), ylab = "f(x)",
      main = "Beta Density Functions")
curve(dbeta(x,3,3), 0, 1, col = "red", add=TRUE)
curve(dbeta(x,5,3), 0, 1, col = "green", add=TRUE)
```


## 4.11  Mixed Distributions  

```{r}
f <- function(x){
  x * (1/100) * exp(-x/100)
}
# here is the answer to 4.157 
integrate(f, 0, 200)$value + (1 - pexp(200, 1/100)) * 200
```

## Independece of two random variables

```{r}
# exercise 
f <- function(x,y){
  (choose(4,x) * choose(3,y) * choose(2, 3 - x - y))/choose(9,3)
}
# use this function to generate all the probabilities for the problem.
```


## Example 6.5  

```{r}
n <- 15
u <- runif(n)
G <- function(u, beta){
  -beta * log(1 - u)
}
G(u, 1)
```

## CH 7, Central Limit Theorem:  

```{r}
# Poisson n = 5
data <- replicate(10000,rpois(5,3))
means <- colMeans(data)
hist(means)
mean(means)
sd(means)
length(which(means > 3.6))

# Poisson n = 50
data <- replicate(10000,rpois(50,3))
means <- colMeans(data)
hist(means)
mean(means)
sd(means)
length(which(means > 3.6))

# exponential n = 5
data <- replicate(10000,rexp(5,.333))
means <- colMeans(data)
hist(means)
mean(means)
sd(means)
length(which(means > 3.6))

# exponential n = 50
data <- replicate(10000,rexp(50,.333))
means <- colMeans(data)
hist(means)
mean(means)
sd(means)
length(which(means > 3.6))
```

## Sampling Distributions  

```{r}
# sample mean - exponential n = 5
data <- replicate(10000,rexp(5,.333))
means <- colMeans(data)
hist(means)
mean(means)
sd(means)
quantile(means,c(0.025,0.975))

# sample mean - exponential n = 50
data <- replicate(10000,rexp(50,.333))
means <- colMeans(data)
hist(means, freq = FALSE, ylim=c(0,1.0))
curve(dnorm(x,mean=3, sd = 3/sqrt(50)), col="darkblue", lwd=2, add=TRUE, yaxt="n")
mean(means)
sd(means)
quantile(means,c(0.025,0.975))

# sample mean - normal n = 5 mean = 100 sd = 10
data <- replicate(10000,rnorm(5, 100, 10))
means <- colMeans(data)
hist(means, freq = FALSE, ylim=c(0,.1))
curve(dnorm(x,mean=100, sd = 10/sqrt(5)), col="darkblue", lwd=2, add=TRUE, yaxt="n")
mean(means)
sd(means)
quantile(means,c(0.025,0.975))
count <- length(which(means > 100+1.96*10/sqrt(5)))
prob <- count/10000
prob

# sample mean curve on original data
hist(data, freq = FALSE, ylim=c(0,.1))
quantile(data,c(0.025,0.975))
curve(dnorm(x,mean=100, sd = 10/sqrt(5)), col="darkblue", lwd=2, add=TRUE, yaxt="n")


# sample variance - normal n = 5 mean = 100 sd = 10
data <- replicate(10000,rnorm(5,100,10))
vars <- apply(data,2,var)
chi <- vars*(5-1)/100
hist(chi, freq = FALSE, ylim=c(0,.20))
curve(dchisq(x,4),0,25,add=T, col = "red")
mean(chi)
sd(chi)
quantile(chi,c(0.025,0.975))

# sample variance - exponential with mean 3
data <- replicate(10000,rexp(5,1/3))
vars <- apply(data,2,var)
chi <- vars*(5-1)/9
hist(chi, freq = FALSE, ylim=c(0,.20))
curve(dchisq(x,4),0,25,add=T, col = "red")
mean(chi)
sd(chi)
quantile(chi,c(0.025,0.975))

# sample variance - poisson mean 3
data <- replicate(10000,rpois(5,3))
vars <- apply(data,2,var)
chi <- vars*(5-1)/3
hist(chi, freq = FALSE, ylim=c(0,.20))
curve(dchisq(x,4),0,25,add=T, col = "red")
mean(chi)
sd(chi)
quantile(chi,c(0.025,0.975))
```

