---
title: "Work WLS"
author: "Cody Frisby"
date: "March 31, 2016"
output:
  html_document:
    keep_md: yes
  pdf_document: default
---

```{r}
work <- read.table("~/Documents/MATH3710/supervisorWLS/work.txt", header = T)
fit <- lm(y~x, work)
summary(fit)
```

Fit model without weights or transformation.  

```{r, dpi=200, fig.align='center'}
par(mfrow=c(1,2))
plot(fit, which = c(1,2))
```


## Add weights to our data table

```{r}
work$w <- 1/work$x^2
work$yt <- work$y/work$x
work$xt <- 1/work$x
```

Now fit the model with the transformed x & y

```{r}
fit.t <- lm(yt ~ xt,data = work)
summary(fit.t)
```

```{r, dpi=200, fig.align='center'}
par(mfrow=c(1,2))
plot(fit.t, which = c(1,2))
```

Now fit the model with weights

```{r}
# not using yt and xt
fit.w <- lm(y ~ x, weights = work$w, data = work)
summary(fit.w)
```

```{r, dpi=200, fig.align='center'}
par(mfrow=c(1,2))
plot(fit.w, which = c(1,2))
```


```{r}
# Here, we calculate the WLS estimates for B_i
X <- cbind(1, work$x)
Y <- matrix(work$y, nrow = length(work$y))
W <- matrix(diag(work$w), nrow = length(work$w))
B <- solve(t(X)%*%W%*%X) %*% t(X)%*%W%*%Y
B # intercept and slope WLS
coef(fit.w) # same as above
```


```{r}
# WSSE(X)
WSSE.x <- sum(work$w*(work$y-(B[1] + B[2]*work$x))^2)
WSSE.x
WMSE.x <- WSSE.x/(length(work$x) - 2)
WMSE.x
sigma0 <- sqrt(WMSE.x)
sigma0 # sigma calculated "by hand"
summary(fit.w)$sigma; summary(fit.t)$sigma #WLS and transformed sigma
```


```{r}
# C matrix
C <- solve(t(X) %*% W %*% X)
C
```


```{r}
# confidence intervals for B
confint(fit.w) # B1 is Significant
```


### Graybill book example pg 576

```{r}
co <- read.table("~/Documents/MATH3710/datafiles/CARBMON.DAT", header = F)
colnames(co) <- c("y", "x")
co$w <- 1/co$x
# using R lm function with weights argument
fit.w <- lm(y ~ x, data = co, weights = w)
y <- co$y; X <- cbind(1, co$x)
# note : w_i = (1/G(x_i)^2) = 1/x_i
W <- matrix(diag(1/co$x), nrow = length(co$x))
B <- solve(t(X)%*%W%*%X) %*% t(X)%*%W%*%y
B; coef(fit.w) # intercept and slope WLS
C <- solve(t(X) %*% W %*% X)
C
```


```{r}
WSSE.x <- sum(co$w*(co$y-(B[1] + B[2]*co$x))^2)
WMSE.x <- WSSE.x/(length(co$x) - 2)
sigma0 <- sqrt(WMSE.x)
sigma0
summary(fit.w)$sigma
```

### Confidence Intervals "By Hand"
```{r}
# SE of B
SE.b0 <- sigma0*sqrt(C[1,1])
SE.b1 <- sigma0*sqrt(C[2,2])
# calculate confidence intervals
cv0 <- qt(0.95, 11)*SE.b0; cv1 <- qt(0.95, 11)*SE.b1
b0 <- B[1] + c(-cv0, cv0)
b1 <- B[2] + c(-cv1, cv1)
rbind(b0,b1)
confint(fit.w, level = 0.90) #compare to R's built in function
```


Confidence interval for sigma and then Y(300)

```{r}
# alpha = 0.2
chi.cv <- qchisq(c(0.9, 0.1), 11)
sqrt(WSSE.x/chi.cv)*sqrt(300) # note, sqrt(300) is the g(y) value
```

Confidence interval for x = 500  

```{r}
# using R "predict"
predict(fit.w, data.frame(x = 500), interval = "confidence")
predict(fit.w, data.frame(x = 500), se.fit = TRUE)$se.fit #SE
# "By hand"
test <- rbind(1,500)
sigma0*sqrt(t(test)%*%C%*%test)
predict(fit.w, data.frame(x=500)) +
  c(-sigma0*sqrt(t(test)%*%C%*%test)*qt(0.975,11),
    sigma0*sqrt(t(test)%*%C%*%test)*qt(0.975,11))
```


Prediction interval for x = 300, pg. 579 in Graybill book 

```{r}
# "By hand"
test <- rbind(1,300)
sigma0*sqrt(sqrt(300)^2 + t(test)%*%C%*%test)
predict(fit.w, data.frame(x=300)) +
  c(-sigma0*sqrt(sqrt(300)^2 + t(test)%*%C%*%test)*qt(0.975,11),
    sigma0*sqrt(sqrt(300)^2 + t(test)%*%C%*%test)*qt(0.975,11))
```


I'm going to try to go further with this and define *w* as a function of *x* using R code to do all the computations for me.

```{r}
# now to attempt the computation using the books notation
# for thw WSSE(X) using R's functionioal programming.
# wwse is a function that takes an 1xn x argument, 1xn y argument
# and an f argument that is a function of x
wls <- function(x, y, w){
  X <- cbind(1, x) # create matrix for computations
  W <- matrix(diag(w), nrow = length(x)) # weight matrix
  B <- solve(t(X) %*% W %*% X) %*% t(X) %*% W %*% y #Betas
  sse <- sum(w*(y-(B[1] + B[2]*x))^2)
  mse <- sse/(length(x) - 2)
  sigma0 <- sqrt(mse)
  z <- list(betas = B, sse = sse, mse = mse, sigma = sigma0) # weighted sum of squares
  z
}
wls(work$x, work$y, 1/work$x^2)
```
